---
title: "Fixed Effects vs Random Effects: A Terminology Guide"
subtitle: "Reconciling Econometrics and Biostatistics Language"
format:
  html:
    theme: cosmo
    toc: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(nlme)
library(lme4)
library(knitr)
library(kableExtra)

# plm is optional - install if needed for econometrics examples
if (!requireNamespace("plm", quietly = TRUE)) {
  message("Note: plm package not installed. Some econometrics examples will be skipped.")
  has_plm <- FALSE
} else {
  library(plm)
  has_plm <- TRUE
}

set.seed(667)
```

## 1. The Terminology Problem

::: {.callout-warning}
## Terminology Alert

**"Random effects" and "fixed effects" mean completely different things in biostatistics vs econometrics!** This is one of the most confusing terminology conflicts in statistics. When reading papers or textbooks, always check which framework the authors are using.
:::

If you are coming from a biostatistics background and encounter econometrics literature (or vice versa), you may find yourself completely confused. The same terms refer to different concepts, different estimators, and different assumptions.

This handout provides a translation guide between the two frameworks and explains when each approach is appropriate.

### The Core Confusion

| When someone says... | A biostatistician thinks... | An econometrician thinks... |
|---------------------|---------------------------|----------------------------|
| "Random effects model" | Mixed model with random intercepts/slopes | Swamy-Arora GLS estimator assuming no endogeneity |
| "Fixed effects model" | Model with only fixed (population-level) parameters | Within-transformation estimator that removes time-invariant confounding |
| "Fixed effects" | The $\beta$ coefficients in a mixed model | Individual-specific intercepts (nuisance parameters) |
| "Random effects" | The $b_i$ deviations from the population mean | An estimation approach, not model components |

## 2. Translation Table

This is the critical reference table for translating between frameworks.

```{r}
#| label: translation-table
#| echo: false

translation <- tibble(
  Term = c(
    "Random effects",
    "Fixed effects",
    "Mixed model",
    "RE model",
    "FE model",
    "Random intercept",
    "Between estimator",
    "Within estimator"
  ),
  `Biostatistics (nlme/lme4)` = c(
    "Subject-specific deviations $b_i$ from population mean",
    "Population-level parameters $\\beta$ (the coefficients of interest)",
    "Model with both fixed $\\beta$ and random $b_i$ components",
    "Model with random intercepts and/or slopes",
    "(Not commonly used this way)",
    "Subject-specific baseline: $\\beta_0 + b_{0i}$",
    "(Not commonly used)",
    "(Not commonly used)"
  ),
  `Econometrics (plm)` = c(
    "Swamy-Arora GLS estimator assuming $\\text{Cor}(X_{it}, \\alpha_i) = 0$",
    "Within-transformation (demeaning) estimator",
    "(Not used in this way)",
    "Estimation assuming no endogeneity",
    "Within estimator that removes time-invariant confounding",
    "Individual effect $\\alpha_i$ (may be correlated with covariates)",
    "OLS on subject means $\\bar{Y}_i$ on $\\bar{X}_i$",
    "OLS on demeaned data: $(Y_{it} - \\bar{Y}_i)$ on $(X_{it} - \\bar{X}_i)$"
  )
)

translation |>
  kable(escape = FALSE, caption = "Translation Between Frameworks") |>
  kable_styling(bootstrap_options = c("striped", "hover", "bordered")) |>
  column_spec(1, bold = TRUE, width = "15%") |>
  column_spec(2, width = "42%") |>
  column_spec(3, width = "42%")
```

### Mathematical Notation Comparison

**Biostatistics (Mixed Model) Notation:**
$$Y_{it} = \mathbf{X}_{it}^\top \boldsymbol{\beta} + \mathbf{Z}_{it}^\top \mathbf{b}_i + \epsilon_{it}$$

where:

- $\boldsymbol{\beta}$ = fixed effects (population parameters)
- $\mathbf{b}_i \sim N(\mathbf{0}, \mathbf{D})$ = random effects (subject deviations)
- $\epsilon_{it} \sim N(0, \sigma^2)$ = residual error

**Econometrics (Panel Data) Notation:**
$$Y_{it} = \mathbf{X}_{it}^\top \boldsymbol{\beta} + \alpha_i + \epsilon_{it}$$

where:

- $\boldsymbol{\beta}$ = coefficients of interest
- $\alpha_i$ = individual effect (may or may not be random)
- $\epsilon_{it}$ = idiosyncratic error

The key difference: econometrics focuses on whether $\alpha_i$ is correlated with $\mathbf{X}_{it}$.

## 3. When Each Framework Applies

```{r}
#| label: situation-table
#| echo: false

situations <- tibble(
  Situation = c(
    "Randomized experiment",
    "Observational data with potential confounding",
    "Interest in time-invariant predictors (e.g., sex, race)",
    "Interest in within-subject change",
    "Small number of subjects, many timepoints",
    "Many subjects, few timepoints",
    "Missing data (MAR assumption reasonable)",
    "Unbalanced panel data"
  ),
  `Biostatistics Approach` = c(
    "Mixed model works well",
    "Mixed model + careful covariate adjustment",
    "Mixed model (can estimate these coefficients)",
    "Mixed model or GEE",
    "Mixed model with careful variance specification",
    "Mixed model or GEE",
    "Mixed model (likelihood-based)",
    "Mixed model handles naturally"
  ),
  `Econometrics Approach` = c(
    "Either FE or RE works",
    "FE estimator preferred (removes confounding)",
    "RE estimator only (FE eliminates these)",
    "FE estimator (focuses on within variation)",
    "FE may have many parameters",
    "Either approach works",
    "Complete cases or specific methods needed",
    "Both handle unbalanced data"
  )
)

situations |>
  kable(caption = "When Each Framework Applies") |>
  kable_styling(bootstrap_options = c("striped", "hover", "bordered")) |>
  column_spec(1, bold = TRUE, width = "30%") |>
  column_spec(2, width = "35%") |>
  column_spec(3, width = "35%")
```

## 4. The Endogeneity Issue

The fundamental question in econometrics is: **Is there endogeneity?**

### What Is Endogeneity?

Endogeneity occurs when covariates are correlated with the individual-specific effect:

$$\text{Cor}(X_{it}, \alpha_i) \neq 0$$

This can happen due to:

1. **Omitted variable bias**: An unobserved time-invariant factor affects both $X$ and $Y$
2. **Selection**: Subjects self-select into treatment based on unobserved characteristics
3. **Reverse causality**: Past outcomes affect current covariates

### Why It Matters

::: {.callout-important}
## Key Insight

If endogeneity exists, the RE estimator (econometrics sense) is **biased and inconsistent**. The biostatistics mixed model is also biased under these conditions, though this is less emphasized in that literature.
:::

**Example**: Studying the effect of job training on wages

- $Y_{it}$ = wages for person $i$ at time $t$
- $X_{it}$ = 1 if person received job training
- $\alpha_i$ = unobserved ability

If more able workers are more likely to seek training, then $\text{Cor}(X_{it}, \alpha_i) > 0$, and a naive analysis will overestimate the training effect.

### How FE Solves the Problem

The FE (within) estimator uses demeaning to remove $\alpha_i$:

$$Y_{it} - \bar{Y}_i = (X_{it} - \bar{X}_i)^\top \beta + (\epsilon_{it} - \bar{\epsilon}_i)$$

Since $\alpha_i$ is constant over time, it cancels out! This removes all time-invariant confounding.

**The cost**: We cannot estimate the effect of time-invariant predictors (they also get differenced out).

### The Hausman Test

The Hausman test compares FE and RE estimates:

- $H_0$: RE is consistent (no endogeneity)
- $H_1$: RE is inconsistent (endogeneity present)

If the test rejects, use FE. If it fails to reject, RE is more efficient.

```{r}
#| label: hausman-example
#| eval: true

# We'll demonstrate with a simulated example in Section 8
# The test statistic compares:
# - FE estimate (consistent under both H0 and H1)
# - RE estimate (efficient under H0, inconsistent under H1)
```

## 5. The Mundlak Approach

The Mundlak approach bridges the two frameworks by including subject-level means as covariates.

### The Basic Idea

Instead of choosing between FE and RE, include the cluster means $\bar{X}_i$ as additional predictors:

$$Y_{it} = \beta_w (X_{it} - \bar{X}_i) + \beta_b \bar{X}_i + \alpha_i + \epsilon_{it}$$

Or equivalently:
$$Y_{it} = \beta_w X_{it} + (\beta_b - \beta_w) \bar{X}_i + \alpha_i + \epsilon_{it}$$

where:

- $\beta_w$ = **within effect**: how changes in $X$ within a subject relate to changes in $Y$
- $\beta_b$ = **between effect**: how differences in average $X$ across subjects relate to differences in average $Y$

### Why This Works

1. **Separates within and between variation**: Can estimate both effects
2. **Allows testing for endogeneity**: If $\beta_w = \beta_b$, no endogeneity
3. **Equivalent to FE for within effects**: $\hat{\beta}_w$ matches the FE estimate
4. **Can include time-invariant predictors**: Unlike pure FE

### The Mundlak Test

Test whether $\beta_w = \beta_b$:

- If equal: RE is valid, effects are the same within and between subjects
- If different: Suggests endogeneity; within and between effects differ

## 6. R Package Comparison

```{r}
#| label: package-table
#| echo: false

packages <- tibble(
  Package = c("nlme", "lme4", "plm", "fixest", "lmerTest"),
  Framework = c(
    "Biostatistics mixed models",
    "Biostatistics mixed models",
    "Econometrics panel data",
    "Econometrics panel data",
    "Biostatistics (extends lme4)"
  ),
  `Key Functions` = c(
    "lme() for mixed models with flexible correlation structures",
    "lmer() for linear mixed models, glmer() for GLMMs",
    "plm() for panel models, phtest() for Hausman test",
    "feols() for fast FE estimation with clustering",
    "lmer() with p-values via Satterthwaite or Kenward-Roger"
  ),
  `Strengths` = c(
    "Flexible correlation/variance structures",
    "Fast computation, large datasets",
    "Full panel econometrics toolkit",
    "Very fast, handles many FE",
    "Proper degrees of freedom for mixed models"
  )
)

packages |>
  kable(caption = "R Packages for Longitudinal/Panel Data") |>
  kable_styling(bootstrap_options = c("striped", "hover", "bordered")) |>
  column_spec(1, bold = TRUE, monospace = TRUE)
```

### Code Comparison: Same Model, Different Packages

```{r}
#| label: code-comparison
#| eval: false

# Generate example panel data
data("Grunfeld", package = "plm")  # Classic econometrics dataset

# -------------------------------------------
# BIOSTATISTICS: Random intercept model (nlme)
# -------------------------------------------
library(nlme)
fit_nlme <- lme(inv ~ value + capital,
                random = ~ 1 | firm,
                data = Grunfeld)

# -------------------------------------------
# BIOSTATISTICS: Random intercept model (lme4)
# -------------------------------------------
library(lme4)
fit_lme4 <- lmer(inv ~ value + capital + (1 | firm),
                 data = Grunfeld)

# -------------------------------------------
# ECONOMETRICS: RE estimator (plm)
# -------------------------------------------
library(plm)
pdata <- pdata.frame(Grunfeld, index = c("firm", "year"))
fit_re <- plm(inv ~ value + capital,
              data = pdata,
              model = "random")

# -------------------------------------------
# ECONOMETRICS: FE estimator (plm)
# -------------------------------------------
fit_fe <- plm(inv ~ value + capital,
              data = pdata,
              model = "within")

# -------------------------------------------
# HAUSMAN TEST: Compare FE and RE
# -------------------------------------------
phtest(fit_fe, fit_re)
```

### Coefficient Comparison

When there is **no endogeneity**, the biostatistics RE (from nlme/lme4) and the econometrics RE (from plm) give similar estimates (though not identical due to different variance estimation).

When there **is endogeneity**, all three (nlme, lme4, plm RE) are biased, but plm FE is consistent.

## 7. Decision Guide

```{r}
#| label: decision-flowchart
#| echo: false
#| fig-width: 8
#| fig-height: 6

library(DiagrammeR)

grViz("
digraph decision_tree {
  graph [rankdir=TB, fontname=Helvetica]
  node [fontname=Helvetica, shape=box, style=filled]

  start [label='Is endogeneity\\na concern?', fillcolor='#E8F4F8']

  experimental [label='Experimental data\\n(randomized)?', fillcolor='#FFF3E0']
  observational [label='Observational\\ndata', fillcolor='#FFF3E0']

  use_mixed [label='Use nlme/lme4\\nfreely', fillcolor='#E8F5E9', shape=ellipse]
  hausman [label='Run Hausman\\ntest', fillcolor='#FCE4EC']

  hausman_result [label='Hausman test\\nrejects?', fillcolor='#E8F4F8']

  use_fe [label='Use FE estimator\\nor Mundlak', fillcolor='#FFEBEE', shape=ellipse]
  use_re [label='RE is valid\\n(more efficient)', fillcolor='#E8F5E9', shape=ellipse]

  time_invariant [label='Need effects of\\ntime-invariant\\npredictors?', fillcolor='#E8F4F8']

  mundlak [label='Use Mundlak\\napproach', fillcolor='#FFF9C4', shape=ellipse]
  fe_only [label='Use FE\\n(cannot estimate\\ntime-invariant)', fillcolor='#FFEBEE', shape=ellipse]

  start -> experimental [label='No']
  start -> observational [label='Yes']

  experimental -> use_mixed
  observational -> hausman

  hausman -> hausman_result

  hausman_result -> use_fe [label='Yes']
  hausman_result -> use_re [label='No']

  use_fe -> time_invariant

  time_invariant -> mundlak [label='Yes']
  time_invariant -> fe_only [label='No']
}
")
```

### Decision Summary

| Question | Answer | Action |
|----------|--------|--------|
| Is this an experiment? | Yes | Use nlme/lme4 mixed models freely |
| Is this observational? | Yes | Consider endogeneity |
| Hausman test significant? | Yes | Use FE or Mundlak approach |
| Hausman test non-significant? | Yes | RE is valid and more efficient |
| Need time-invariant predictor effects? | Yes | Use Mundlak (FE eliminates these) |

## 8. Worked Example

Let us analyze the same simulated dataset using both frameworks to see when they agree and when they diverge.

### Simulate Panel Data

```{r}
#| label: simulate-data

set.seed(667)

n_subjects <- 100
n_time <- 5

# Create panel data with potential endogeneity
panel_data <- expand_grid(
  id = 1:n_subjects,
  time = 1:n_time
) |>
  mutate(
    # Unobserved individual effect (ability)
    alpha_i = rep(rnorm(n_subjects, mean = 0, sd = 2), each = n_time),

    # Time-invariant predictor (education years)
    education = rep(sample(12:20, n_subjects, replace = TRUE), each = n_time),

    # Time-varying predictor (job training)
    # KEY: Training is correlated with alpha_i (endogeneity!)
    training = rbinom(n_subjects * n_time, 1,
                      prob = plogis(-1 + 0.3 * rep(rnorm(n_subjects), each = n_time) +
                                    0.2 * alpha_i)),

    # Another time-varying predictor (no endogeneity)
    experience = time + rep(sample(0:10, n_subjects, replace = TRUE), each = n_time),

    # Outcome (wages)
    # True effects: training = 2, education = 0.5, experience = 0.3
    wages = 10 + 2 * training + 0.5 * education + 0.3 * experience +
            alpha_i + rnorm(n_subjects * n_time, sd = 1)
  )

head(panel_data, 10)
```

### Check for Endogeneity

```{r}
#| label: check-endogeneity

# Correlation between training and unobserved effect
panel_data |>
  group_by(id) |>
  summarise(
    mean_training = mean(training),
    alpha = first(alpha_i)
  ) |>
  summarise(
    correlation = cor(mean_training, alpha)
  )
```

We see that training is positively correlated with the unobserved individual effect (ability). This is endogeneity.

### Analysis 1: Biostatistics Mixed Model (lme4)

```{r}
#| label: mixed-model

library(lme4)

fit_mixed <- lmer(wages ~ training + education + experience + (1 | id),
                  data = panel_data)
summary(fit_mixed)
```

**Result**: The training effect is estimated as `r round(fixef(fit_mixed)["training"], 3)`, which is **biased upward** from the true value of 2.0. This is because the mixed model does not account for the correlation between training and the unobserved individual effect.

### Analysis 2: Econometrics Panel Models (plm)

```{r}
#| label: panel-models

library(plm)

# Convert to panel data frame
pdata <- pdata.frame(panel_data, index = c("id", "time"))

# RE estimator
fit_re <- plm(wages ~ training + education + experience,
              data = pdata,
              model = "random")

# FE estimator
fit_fe <- plm(wages ~ training + education + experience,
              data = pdata,
              model = "within")

cat("RE Estimator:\n")
summary(fit_re)$coefficients

cat("\n\nFE (Within) Estimator:\n")
summary(fit_fe)$coefficients
```

**Key observations**:

1. **RE estimate for training**: Similar to mixed model, biased upward
2. **FE estimate for training**: `r round(coef(fit_fe)["training"], 3)`, much closer to true value of 2.0
3. **Education coefficient**: Estimated by RE, but **NOT by FE** (time-invariant, differenced out)

### Analysis 3: Hausman Test

```{r}
#| label: hausman-test

# Hausman test: compare FE and RE
hausman_test <- phtest(fit_fe, fit_re)
hausman_test
```

The Hausman test is **significant** (p < 0.05), indicating that the RE estimator is inconsistent and we should use FE.

### Analysis 4: Mundlak Approach

```{r}
#| label: mundlak

# Add cluster means
panel_data <- panel_data |>
  group_by(id) |>
  mutate(
    training_mean = mean(training),
    experience_mean = mean(experience)
  ) |>
  ungroup()

# Mundlak model using lme4
fit_mundlak <- lmer(wages ~ training + education + experience +
                    training_mean + experience_mean + (1 | id),
                    data = panel_data)
summary(fit_mundlak)
```

**Mundlak results**:

- **Within effect of training** (`training`): `r round(fixef(fit_mundlak)["training"], 3)` -- close to true value!
- **Between effect of training** (`training_mean`): `r round(fixef(fit_mundlak)["training_mean"], 3)` -- different from within effect
- **Education effect**: `r round(fixef(fit_mundlak)["education"], 3)` -- can now estimate this!

The difference between within and between effects for training suggests endogeneity, consistent with our Hausman test.

### Summary Comparison

```{r}
#| label: comparison-table

comparison <- tibble(
  Method = c("True Value", "Mixed Model (lme4)", "RE (plm)", "FE (plm)", "Mundlak"),
  `Training Effect` = c(
    2.0,
    round(fixef(fit_mixed)["training"], 3),
    round(coef(fit_re)["training"], 3),
    round(coef(fit_fe)["training"], 3),
    round(fixef(fit_mundlak)["training"], 3)
  ),
  `Education Effect` = c(
    0.5,
    round(fixef(fit_mixed)["education"], 3),
    round(coef(fit_re)["education"], 3),
    NA,  # FE cannot estimate
    round(fixef(fit_mundlak)["education"], 3)
  ),
  `Experience Effect` = c(
    0.3,
    round(fixef(fit_mixed)["experience"], 3),
    round(coef(fit_re)["experience"], 3),
    round(coef(fit_fe)["experience"], 3),
    round(fixef(fit_mundlak)["experience"], 3)
  ),
  Notes = c(
    "Ground truth",
    "Biased by endogeneity",
    "Biased by endogeneity",
    "Consistent, but loses education",
    "Best of both worlds"
  )
)

comparison |>
  kable(caption = "Comparison of Estimators", digits = 3) |>
  kable_styling(bootstrap_options = c("striped", "hover", "bordered")) |>
  column_spec(1, bold = TRUE) |>
  row_spec(1, background = "#E8F5E9") |>
  row_spec(4:5, background = "#FFF9C4")
```

### Visualizing Within vs Between Variation

```{r}
#| label: within-between-plot
#| fig-width: 10
#| fig-height: 5

# Show the within vs between variation
p1 <- panel_data |>
  ggplot(aes(x = training_mean, y = wages, group = id)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(
    title = "Between Variation",
    subtitle = "Subject means (confounded)",
    x = "Mean Training (by subject)",
    y = "Wages"
  ) +
  theme_minimal()

p2 <- panel_data |>
  mutate(
    training_demeaned = training - training_mean,
    wages_demeaned = wages - mean(wages)  # simplified for visualization
  ) |>
  ggplot(aes(x = training_demeaned, y = wages_demeaned)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(
    title = "Within Variation",
    subtitle = "Demeaned (FE uses this)",
    x = "Training - Subject Mean",
    y = "Wages (demeaned)"
  ) +
  theme_minimal()

library(patchwork)
p1 + p2
```

## Key Takeaways

::: {.callout-tip}
## Summary

1. **Check the framework** when reading papers: biostatistics "RE" $\neq$ econometrics "RE"

2. **Endogeneity matters**: If time-invariant confounders exist, use FE or Mundlak

3. **Hausman test** helps decide: significant = use FE; non-significant = RE is efficient

4. **Mundlak approach** gives both within and between effects, and allows time-invariant predictors

5. **Experimental data** avoids endogeneity by design; mixed models work well
:::

::: {.callout-warning}
## When Reading Papers, Always Check Which Framework the Authors Are Using!

The same terms mean different things. Look for clues:

- "Within estimator" or "demeaning" = econometrics FE
- "Random intercept model" = biostatistics mixed model
- Hausman test mentioned = econometrics framework
- `lme4` or `nlme` used = biostatistics framework
- `plm` or `fixest` used = econometrics framework
:::

## Session Information

```{r}
#| label: session-info

sessionInfo()
```
