---
title: "GEE Complete Guide"
subtitle: "Generalized Estimating Equations for Longitudinal Data"
format:
  html:
    theme: cosmo
    toc: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(geepack)
library(lme4)
library(MASS)
```

## 1. What is GEE?

Generalized Estimating Equations (GEE) provide a **population-averaged (marginal)** approach for analyzing correlated data, particularly longitudinal and clustered outcomes.

### Key Features

| Feature | Description |
|---------|-------------|
| **Target** | Population-averaged (marginal) effects |
| **Distribution** | Does not require full distribution specification |
| **Robustness** | Robust to working correlation misspecification (with sandwich SE) |
| **Estimation** | Quasi-likelihood approach |
| **Standard Errors** | Uses sandwich (robust) estimator |

### Core Idea

GEE answers: *"How does the average response in the population change with covariates?"*

Unlike GLMMs, GEE does not model individual-specific trajectories. Instead, it directly estimates the marginal mean:

$$E(Y_{ij} | X_{ij}) = g^{-1}(X_{ij}^T \beta)$$

where $g$ is the link function (e.g., logit for binary outcomes).

::: {.callout-important}
## The Sandwich Estimator is Key

GEE's robustness comes from the **sandwich (robust) standard errors**, which remain valid even if the working correlation structure is misspecified. Always use robust SEs in practice!
:::


## 2. GEE vs GLMM Comparison

| Aspect | GEE | GLMM |
|--------|-----|------|
| **Target** | Population average | Subject-specific |
| **Interpretation** | Marginal effects | Conditional on random effects |
| **Requires distribution** | No (quasi-likelihood) | Yes (full likelihood) |
| **Random effects** | No | Yes |
| **Robust to correlation misspecification** | Yes (with sandwich SE) | No |
| **Valid under MAR** | No (need IPW-GEE) | Yes |
| **Coefficient magnitude (binary)** | Smaller | Larger |
| **Model comparison** | QIC | AIC/BIC/LRT |

### When to Use Each

| Scenario | Recommended Approach |
|----------|---------------------|
| Interest in population-level inference | **GEE** |
| Interest in individual trajectories | **GLMM** |
| Few clusters (< 30) | **GLMM** (or GEE with small-sample correction) |
| Correlation structure uncertain | **GEE** (robust to misspecification) |
| Missing data (MAR mechanism) | **GLMM** or **IPW-GEE** |
| Simple, robust analysis | **GEE** |

### Coefficient Interpretation Example (Binary Outcome)

For a treatment effect with binary outcome (logit link):

- **GEE**: "The odds of the outcome in the treated group are $e^{\beta}$ times the odds in the control group, on average in the population."

- **GLMM**: "For a given subject, the odds of the outcome when treated are $e^{\beta}$ times the odds when not treated, holding random effects constant."

::: {.callout-note}
## Marginal vs Conditional

For non-linear models (e.g., logistic), marginal (GEE) coefficients are attenuated toward zero compared to conditional (GLMM) coefficients. This is not a bias---it reflects different targets of inference.
:::


## 3. Working Correlation Structures

GEE uses a **working correlation** structure to improve efficiency. The key insight: even if misspecified, sandwich SEs remain valid.

| Structure | Assumption | When to Use | `geepack` syntax |
|-----------|------------|-------------|------------------|
| **Independence** | $\text{Corr}(Y_{ij}, Y_{ik}) = 0$ | Baseline; conservative | `corstr = "independence"` |
| **Exchangeable** | $\text{Corr}(Y_{ij}, Y_{ik}) = \rho$ for all $j \neq k$ | Clustered data; similar within-cluster correlation | `corstr = "exchangeable"` |
| **AR(1)** | $\text{Corr}(Y_{ij}, Y_{ik}) = \rho^{|j-k|}$ | Equally-spaced longitudinal data; correlation decays | `corstr = "ar1"` |
| **Unstructured** | All correlations freely estimated | Small clusters; exploratory | `corstr = "unstructured"` |

### Correlation Structure Formulas

**Exchangeable (Compound Symmetry)**:
$$R_{jk} = \begin{cases} 1 & j = k \\ \rho & j \neq k \end{cases}$$

**AR(1)**:
$$R_{jk} = \rho^{|j-k|}$$

**Unstructured**:
$$R_{jk} = \rho_{jk} \text{ (each pair estimated separately)}$$

### Practical Guidance

1. **Start with exchangeable** for clustered data or longitudinal data with moderate follow-up
2. **Use AR(1)** for longitudinal data with time-ordered measurements
3. **Independence** gives valid (but inefficient) inference
4. **Unstructured** requires many parameters---use only with small cluster sizes

::: {.callout-tip}
## Robustness Matters More Than Efficiency

Choosing the "correct" working correlation improves efficiency but is not required for valid inference. With robust SEs, any working correlation gives valid hypothesis tests and confidence intervals.
:::


## 4. The Sandwich Estimator

### Why Standard (Model-Based) SEs Fail

Model-based SEs assume the working correlation is correctly specified. If wrong, they can be severely biased (either direction).

### The Sandwich Formula

The robust variance estimator has the "sandwich" form:

$$\widehat{\text{Var}}(\hat{\beta}) = B^{-1} M B^{-1}$$

where:

- $B$ = model-based information matrix ("bread")
- $M$ = empirical covariance of score contributions ("meat")

### Key Properties

| Property | Explanation |
|----------|-------------|
| **Consistent** | Converges to true variance as $n \to \infty$ |
| **Robust** | Valid even with misspecified working correlation |
| **Cluster-robust** | Accounts for within-cluster dependence |

### The Trade-off

::: {.callout-warning}
## Sample Size Requirement

The sandwich estimator requires **enough clusters** for reliable performance:

- Rule of thumb: $n \geq 30-50$ clusters
- With fewer clusters, sandwich SEs tend to be **biased downward** (anti-conservative)
- Use small-sample corrections when $n < 40$
:::


## 5. Small Sample Corrections

### The Problem

With few clusters, the sandwich estimator underestimates variance, leading to:

- Inflated Type I error rates
- Confidence intervals that are too narrow

### Solutions

| Correction | Description | When to Use |
|------------|-------------|-------------|
| **Bias-corrected (BC)** | Adjusts for finite-sample bias | $n = 20-40$ clusters |
| **Jackknife** | Leave-one-cluster-out variance | $n = 15-40$ clusters |
| **Bootstrap** | Cluster bootstrap | Flexible; any $n$ |
| **Fay's correction** | Modified BC | Very small $n$ |

### R Implementation with `geepack`

```{r}
#| eval: false

# Standard fit
fit <- geeglm(y ~ x, id = id, data = dat,
              family = binomial, corstr = "exchangeable")

# Extract robust SEs (default)
summary(fit)$coefficients[, "Std.err"]

# For small-sample corrections, use additional packages:
# - geesmv: small-sample variance corrections
# - clubSandwich: cluster-robust SEs with corrections
```

### Using `clubSandwich` for Small Samples

```{r}
#| eval: false

library(clubSandwich)

# Fit GEE
fit <- geeglm(y ~ x, id = id, data = dat,
              family = binomial, corstr = "exchangeable")

# CR2 correction (recommended for small samples)
coef_test(fit, vcov = "CR2", cluster = dat$id)
```


## 6. Complete R Example

### Setup: Epilepsy Data Pattern

```{r}
set.seed(667)

# Simulate epilepsy-like data
n_subjects <- 60
n_visits <- 4

# Create data structure
dat <- data.frame(
  id = rep(1:n_subjects, each = n_visits),
  visit = rep(0:(n_visits-1), n_subjects),
  trt = rep(rbinom(n_subjects, 1, 0.5), each = n_visits),
  age = rep(rnorm(n_subjects, 30, 10), each = n_visits)
)

# Generate subject-specific random effects
subject_re <- rnorm(n_subjects, 0, 0.8)

# Generate correlated binary outcomes (seizure-free: yes/no)
dat$eta <- -1 + 0.5 * dat$trt + 0.3 * dat$visit - 0.02 * dat$age +
           subject_re[dat$id]
dat$prob <- plogis(dat$eta)
dat$seizure_free <- rbinom(nrow(dat), 1, dat$prob)

# Convert id to factor
dat$id <- factor(dat$id)

head(dat)
```

### Fit GEE with Exchangeable Correlation

```{r}
# Fit GEE
fit_gee <- geeglm(seizure_free ~ trt + visit + age,
                  id = id,
                  data = dat,
                  family = binomial(link = "logit"),
                  corstr = "exchangeable")

summary(fit_gee)
```

### Extract and Interpret Results

```{r}
# Extract coefficients
coefs <- summary(fit_gee)$coefficients
print(coefs)

# Odds ratios with 95% CI
or_results <- data.frame(
  Variable = rownames(coefs),
  OR = exp(coefs[, "Estimate"]),
  Lower = exp(coefs[, "Estimate"] - 1.96 * coefs[, "Std.err"]),
  Upper = exp(coefs[, "Estimate"] + 1.96 * coefs[, "Std.err"]),
  P_value = coefs[, "Pr(>|W|)"]
)
print(or_results, digits = 3)
```

### Working Correlation Estimate

```{r}
# Extract estimated correlation
fit_gee$geese$alpha
```

### Compare to GLMM

```{r}
# Fit GLMM for comparison
fit_glmm <- glmer(seizure_free ~ trt + visit + age + (1 | id),
                  data = dat, family = binomial)

# Compare coefficients
comparison <- data.frame(
  Variable = names(fixef(fit_glmm)),
  GEE = coef(fit_gee),
  GLMM = fixef(fit_glmm)
)
comparison$Ratio <- comparison$GLMM / comparison$GEE
print(comparison, digits = 3)
```

::: {.callout-note}
## Coefficient Comparison

Notice that GLMM coefficients are larger in magnitude than GEE coefficients. This is expected for binary outcomes---marginal effects are attenuated compared to conditional effects.
:::

### Compare Correlation Structures

```{r}
# Independence
fit_ind <- geeglm(seizure_free ~ trt + visit + age,
                  id = id, data = dat, family = binomial,
                  corstr = "independence")

# Exchangeable (already fit above)

# AR(1)
fit_ar1 <- geeglm(seizure_free ~ trt + visit + age,
                  id = id, data = dat, family = binomial,
                  corstr = "ar1")

# Unstructured
fit_un <- geeglm(seizure_free ~ trt + visit + age,
                 id = id, data = dat, family = binomial,
                 corstr = "unstructured")

# Compare QIC (not AIC!)
qic_compare <- data.frame(
  Structure = c("Independence", "Exchangeable", "AR(1)", "Unstructured"),
  QIC = c(QIC(fit_ind)[1], QIC(fit_gee)[1], QIC(fit_ar1)[1], QIC(fit_un)[1])
)
print(qic_compare)
```


## 7. GEE Diagnostics

### Residual Analysis

```{r}
# Pearson residuals
dat$pearson_resid <- residuals(fit_gee, type = "pearson")

# Plot residuals vs fitted
plot(fitted(fit_gee), dat$pearson_resid,
     xlab = "Fitted probabilities", ylab = "Pearson residuals",
     main = "GEE Residuals vs Fitted")
abline(h = 0, lty = 2, col = "red")
```

### Checking Working Correlation

```{r}
# Compare coefficient estimates across structures
coef_compare <- data.frame(
  Independence = coef(fit_ind),
  Exchangeable = coef(fit_gee),
  AR1 = coef(fit_ar1),
  Unstructured = coef(fit_un)
)
round(coef_compare, 3)
```

::: {.callout-tip}
## Stability Across Structures

If coefficient estimates are similar across working correlation structures, this suggests robustness. Large differences may indicate model misspecification or influential observations.
:::

### QIC for Model Selection

**Use QIC (Quasi-likelihood under Independence Criterion), NOT AIC!**

QIC is the appropriate analog of AIC for GEE:

$$\text{QIC} = -2Q(\hat{\beta}) + 2p$$

where $Q$ is the quasi-likelihood and $p$ is the number of parameters.

```{r}
# QIC comparison
cat("QIC values:\n")
cat("Independence:", QIC(fit_ind)[1], "\n")
cat("Exchangeable:", QIC(fit_gee)[1], "\n")
cat("AR(1):", QIC(fit_ar1)[1], "\n")
cat("Unstructured:", QIC(fit_un)[1], "\n")
```


## 8. IPW-GEE for MAR

### The Problem

Standard GEE is valid only under **MCAR** (Missing Completely at Random). Under **MAR** (Missing at Random), GEE estimates can be biased.

### Solution: Inverse Probability Weighting

IPW-GEE weights observations by the inverse probability of being observed:

$$w_{ij} = \frac{1}{P(\text{observed}_{ij} | \text{covariates})}$$

### Implementation Sketch

```{r}
#| eval: false

# Step 1: Model missingness
miss_model <- glm(observed ~ lag_y + x + time,
                  data = dat, family = binomial)

# Step 2: Calculate weights
dat$prob_obs <- predict(miss_model, type = "response")
dat$ipw <- 1 / dat$prob_obs

# Step 3: Fit weighted GEE
fit_ipw <- geeglm(y ~ x + time, id = id, data = dat[dat$observed == 1, ],
                  family = binomial, corstr = "exchangeable",
                  weights = ipw)
```

::: {.callout-warning}
## IPW-GEE Considerations

- Requires correct model for missingness mechanism
- Extreme weights can inflate variance
- Often combined with weight stabilization or trimming
:::


## 9. Common Mistakes

### Mistake 1: Using Model-Based SEs

```{r}
#| eval: false

# WRONG: naive/model-based SEs
# (geepack uses robust by default, but other packages may not)

# CORRECT: Always use robust (sandwich) SEs
summary(fit_gee)$coefficients[, "Std.err"]  # These ARE robust in geepack
```

### Mistake 2: Using AIC Instead of QIC

```{r}
#| eval: false

# WRONG
AIC(fit_gee)  # Not appropriate for GEE

# CORRECT
QIC(fit_gee)  # Use QIC for quasi-likelihood models
```

### Mistake 3: Too Few Clusters

| Number of Clusters | Recommendation |
|--------------------|----------------|
| $n < 20$ | Consider GLMM instead |
| $n = 20-40$ | Use small-sample corrections |
| $n > 40$ | Standard GEE likely okay |

### Mistake 4: Subject-Specific Interpretation

```{r}
#| eval: false

# WRONG interpretation:
# "For an individual patient, treatment increases odds by..."

# CORRECT interpretation:
# "In the population, treatment is associated with X times higher odds of..."
```

### Mistake 5: Ignoring MAR

```{r}
#| eval: false

# WRONG: Standard GEE with dropout
fit <- geeglm(y ~ trt + time, id = id, data = dat, ...)  # Biased under MAR

# CORRECT: Use IPW-GEE or GLMM
```


## 10. Quick Reference

### Essential Code

| Task | Code |
|------|------|
| Fit GEE | `geeglm(y ~ x, id = id, family = binomial, corstr = "exchangeable", data = dat)` |
| Get robust SEs | `summary(fit)$coefficients[, "Std.err"]` |
| Odds ratios | `exp(coef(fit))` |
| 95% CI | `exp(coef(fit) +/- 1.96 * SE)` |
| Working correlation | `fit$geese$alpha` |
| Model comparison | `QIC(fit1); QIC(fit2)` |

### Correlation Structure Quick Guide

| Data Pattern | Recommended Structure |
|--------------|----------------------|
| Clustered (e.g., patients in clinics) | `exchangeable` |
| Longitudinal, equal spacing | `ar1` |
| Longitudinal, irregular spacing | `exchangeable` or `independence` |
| Few time points per subject | `unstructured` |
| Uncertain | `independence` (conservative) |

### Decision Flowchart

```
Population-level inference? --> Yes --> GEE
                           --> No  --> Consider GLMM

Many clusters (n > 40)?    --> Yes --> Standard GEE
                           --> No  --> Small-sample correction or GLMM

Missing data pattern?      --> MCAR --> Standard GEE
                           --> MAR  --> IPW-GEE or GLMM
                           --> MNAR --> Sensitivity analysis needed
```

### Package Reference

```{r}
#| eval: false

# Primary package
library(geepack)    # geeglm(), QIC()

# Small-sample corrections
library(clubSandwich)  # coef_test() with CR2

# Comparison with GLMM
library(lme4)       # glmer()

# IPW-GEE
library(ipw)        # ipwpoint(), ipwtm()
```

::: {.callout-tip}
## Summary

1. **GEE gives population-averaged effects** with robust inference
2. **Always use sandwich (robust) SEs**---they protect against correlation misspecification
3. **Use QIC, not AIC** for model comparison
4. **Need 30+ clusters** for reliable sandwich SEs (or use corrections)
5. **Standard GEE assumes MCAR**---use IPW-GEE or GLMM under MAR
:::
