---
title: "BIOS 667 — Estimation & Inference Quick Reference"
subtitle: "ML, REML, GLS, and Hypothesis Testing (Ch. 4)"
format:
  html:
    theme: cosmo
    toc: true
    embed-resources: true
  pdf:
    number-sections: false
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(nlme)
library(lme4)
```

## 1. Estimation Methods Comparison

| Method | When to Use | Pros | Cons | R Code |
|--------|-------------|------|------|--------|
| **OLS** | Baseline comparison; independent errors | Simple, familiar | Ignores correlation; SEs wrong | `lm(y ~ x)` |
| **GLS** | Known or estimated $\Sigma$ | Correct SEs; efficient | Requires specifying structure | `gls(y ~ x, correlation = ...)` |
| **ML** | Comparing models with different **fixed effects** | LRT valid for fixed effects | Variance estimates biased downward | `gls(..., method = "ML")` |
| **REML** | Estimating **variance components** | Less biased variance estimates | Cannot compare models with different fixed effects | `gls(..., method = "REML")` |

**Key insight**: OLS point estimates for $\beta$ are unbiased even with correlated errors, but the standard errors are wrong. GLS gives both correct estimates AND correct SEs.


## 2. ML vs REML Decision Guide

### When to Use Each

| Goal | Method | Reason |
|------|--------|--------|
| Estimate variance components | **REML** | Less biased (accounts for estimating fixed effects) |
| Compare models with same fixed effects, different covariance | Either (REML preferred) | Variance comparison only |
| Compare models with **different fixed effects** | **ML required** | REML likelihoods not comparable |
| Final model reporting | **REML** | Better variance estimates |

### The REML Gotcha

::: {.callout-warning}
## Critical: Use ML for LRT on Fixed Effects

REML likelihoods are **not comparable** across models with different fixed effects because REML integrates out the fixed effects. If you use REML for an LRT comparing fixed effects, your test is invalid!
:::

### Code Pattern

```{r}
#| eval: false

# Step 1: Fit models with ML for comparison
fit_full <- gls(y ~ x1 + x2, data = dat,
                correlation = corAR1(form = ~ time | id),
                method = "ML")

fit_reduced <- gls(y ~ x1, data = dat,
                   correlation = corAR1(form = ~ time | id),
                   method = "ML")

# Step 2: LRT for fixed effects (valid with ML)
anova(fit_reduced, fit_full)

# Step 3: Refit final model with REML for inference
fit_final <- gls(y ~ x1 + x2, data = dat,
                 correlation = corAR1(form = ~ time | id),
                 method = "REML")
```


## 3. Hypothesis Testing Reference

| Test | Formula | When to Use | R Function |
|------|---------|-------------|------------|
| **Wald** | $W = \frac{(\hat{\beta} - \beta_0)^2}{\widehat{\text{Var}}(\hat{\beta})} \sim \chi^2_1$ | Quick tests; large samples | `summary()` t-values |
| **LRT** | $\Lambda = -2[\ell(\theta_0) - \ell(\hat{\theta})] \sim \chi^2_{df}$ | Nested models; use **ML** for fixed effects | `anova(fit1, fit2)` |
| **Score** | Based on gradient at $\theta_0$ | Rarely used directly | Not standard in R |

### Wald vs LRT

- **Wald**: Uses only the full model fit; can be unreliable far from null
- **LRT**: Requires fitting both models; generally more reliable
- **For variance components**: LRT at boundary (testing $\sigma^2 = 0$) has non-standard distribution; consider mixture of $\chi^2$

### LRT Example

```{r}
#| eval: false

# Testing fixed effects (MUST use ML)
fit1_ml <- gls(y ~ x1, data = dat, correlation = corCS(form = ~ 1 | id),
               method = "ML")
fit2_ml <- gls(y ~ x1 + x2, data = dat, correlation = corCS(form = ~ 1 | id),
               method = "ML")

anova(fit1_ml, fit2_ml)  # Valid LRT

# Testing covariance structures (can use REML, same fixed effects)
fit_cs <- gls(y ~ x1 + x2, data = dat, correlation = corCS(form = ~ 1 | id),
              method = "REML")
fit_ar1 <- gls(y ~ x1 + x2, data = dat, correlation = corAR1(form = ~ time | id),
               method = "REML")

# Compare via AIC/BIC (not nested, so LRT inappropriate)
AIC(fit_cs, fit_ar1)
```


## 4. Covariance Structure Selection

| Structure | Parameters | Correlation Pattern | When to Use |
|-----------|------------|---------------------|-------------|
| **Independence** | 1 ($\sigma^2$) | $\text{Corr}(Y_i, Y_j) = 0$ | Baseline; rarely appropriate for longitudinal |
| **Compound Symmetry (CS)** | 2 ($\sigma^2, \rho$) | $\text{Corr}(Y_i, Y_j) = \rho$ for all $i \neq j$ | Equal spacing; exchangeable observations |
| **AR(1)** | 2 ($\sigma^2, \rho$) | $\text{Corr}(Y_i, Y_j) = \rho^{|i-j|}$ | Equally spaced; correlation decays with lag |
| **Unstructured (UN)** | $n_i(n_i+1)/2$ | All correlations estimated freely | Small $n_i$; exploratory; gold standard |

### Parameter Count for Unstructured

With $n_i$ observations per subject:

- Variances: $n_i$
- Covariances: $\binom{n_i}{2} = \frac{n_i(n_i-1)}{2}$
- **Total**: $\frac{n_i(n_i+1)}{2}$

Example: 5 time points $\rightarrow$ $\frac{5 \times 6}{2} = 15$ parameters

### Selection Strategy

1. Start with unstructured (if feasible) as reference
2. Compare simpler structures via AIC/BIC
3. Consider scientific plausibility (does correlation decay with time?)
4. Parsimony matters: fewer parameters = more stable estimates


## 5. R Code Examples

### Setup: Simulated Longitudinal Data

```{r}
# Simulate data
set.seed(667)
n_subj <- 50
n_time <- 4

dat <- data.frame(
  id = rep(1:n_subj, each = n_time),
  time = rep(1:n_time, n_subj),
  x = rnorm(n_subj * n_time)
)

# Add correlated errors (AR1 structure)
library(MASS)
ar1_cor <- function(n, rho) {
  mat <- rho^abs(outer(1:n, 1:n, "-"))
  return(mat)
}

errors <- do.call(rbind, lapply(1:n_subj, function(i) {
  mvrnorm(1, mu = rep(0, n_time), Sigma = 4 * ar1_cor(n_time, 0.6))
}))

dat$y <- 2 + 0.5 * dat$x + as.vector(t(errors))
dat$id <- factor(dat$id)
```

### Fitting GLS with Different Structures

```{r}
# Independence (equivalent to OLS)
fit_ind <- gls(y ~ x, data = dat, method = "REML")

# Compound Symmetry
fit_cs <- gls(y ~ x, data = dat,
              correlation = corCompSymm(form = ~ 1 | id),
              method = "REML")

# AR(1)
fit_ar1 <- gls(y ~ x, data = dat,
               correlation = corAR1(form = ~ time | id),
               method = "REML")

# Compare fits
AIC(fit_ind, fit_cs, fit_ar1)
```

### Extracting and Comparing Standard Errors

```{r}
# Extract coefficients and SEs
extract_results <- function(fit, name) {
  data.frame(
    Model = name,
    Estimate = coef(fit)["x"],
    SE = sqrt(vcov(fit)["x", "x"])
  )
}

results <- rbind(
  extract_results(fit_ind, "Independence"),
  extract_results(fit_cs, "Compound Symmetry"),
  extract_results(fit_ar1, "AR(1)")
)
print(results, row.names = FALSE)
```

### LRT for Nested Models (Fixed Effects)

```{r}
# Fit with ML for valid LRT
fit_reduced_ml <- gls(y ~ 1, data = dat,
                      correlation = corAR1(form = ~ time | id),
                      method = "ML")

fit_full_ml <- gls(y ~ x, data = dat,
                   correlation = corAR1(form = ~ time | id),
                   method = "ML")

# LRT
anova(fit_reduced_ml, fit_full_ml)
```

### Examining Estimated Correlation

```{r}
# Extract correlation parameter from AR(1) fit
fit_ar1$modelStruct$corStruct

# Get the estimated correlation matrix for first subject
getVarCov(fit_ar1, individual = 1, type = "marginal")
```


## Quick Reference Card

::: {.callout-tip}
## Estimation & Testing Checklist

1. **Comparing fixed effects?** $\rightarrow$ Use `method = "ML"`
2. **Final inference on variances?** $\rightarrow$ Use `method = "REML"`
3. **Nested models?** $\rightarrow$ LRT via `anova()`
4. **Non-nested covariance structures?** $\rightarrow$ AIC/BIC
5. **Choosing structure?** $\rightarrow$ Start UN, simplify with AIC
:::

| I want to... | Use... |
|--------------|--------|
| Test if a covariate is significant | Wald (quick) or LRT with ML (rigorous) |
| Compare AR(1) vs CS (same fixed effects) | AIC/BIC or LRT with REML |
| Get best variance estimates | REML |
| Test $H_0: \beta_1 = 0$ via LRT | ML for both models, then `anova()` |
