---
title: "Ordinal Regression Coefficient Interpretation"
subtitle: "Cumulative Logit Models and Proportional Odds"
format:
  html:
    theme: cosmo
    toc: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(MASS)
library(ordinal)
library(emmeans)
library(ggplot2)
```

## 1. What is Ordinal Regression?

Ordinal regression is used when the outcome variable has **ordered categories** that have a natural ranking but the distances between categories are not necessarily equal.

### Examples of Ordinal Outcomes

| Domain | Categories |
|--------|------------|
| **Pain severity** | None < Mild < Moderate < Severe |
| **Disease stage** | Stage I < Stage II < Stage III < Stage IV |
| **Agreement scale** | Strongly Disagree < Disagree < Neutral < Agree < Strongly Agree |
| **Functional status** | Independent < Some assistance < Dependent |
| **Treatment response** | No response < Partial response < Complete response |

### The Cumulative Logit Model

The most common ordinal regression model is the **cumulative logit model** (also called proportional odds model). It models the probability of being at or below each category.

**Link function:**
$$\text{logit}(P(Y \leq j)) = \log\left(\frac{P(Y \leq j)}{P(Y > j)}\right)$$

This is the log-odds of being in category $j$ or lower versus being in a higher category.

### Key Feature: Single Coefficient

The proportional odds model uses a **single coefficient** for each predictor across all cut points. This assumes the effect of the predictor is the same regardless of where you "cut" the outcome scale.

::: {.callout-important}
## Why Ordinal, Not Multinomial?

Ordinal regression **uses the ordering information** in your outcome. Multinomial regression ignores this ordering, treating categories as unrelated. Ordinal regression is more efficient and interpretable when the outcome is truly ordered.
:::


## 2. The Model

### Mathematical Formulation

For an outcome $Y$ with $J$ ordered categories (1, 2, ..., J), the cumulative logit model is:

$$\text{logit}(P(Y \leq j | \mathbf{X})) = \alpha_j - \boldsymbol{\beta}^T\mathbf{X}$$

for $j = 1, 2, \ldots, J-1$

where:

| Component | Description |
|-----------|-------------|
| $\alpha_j$ | Threshold (intercept) for cumulative probability up to category $j$ |
| $\boldsymbol{\beta}$ | Vector of regression coefficients |
| $\mathbf{X}$ | Vector of covariates |

### Understanding the Thresholds

There are $J-1$ thresholds for $J$ categories:

- $\alpha_1$: threshold between category 1 and categories 2-J
- $\alpha_2$: threshold between categories 1-2 and categories 3-J
- $\vdots$
- $\alpha_{J-1}$: threshold between categories 1 to J-1 and category J

The thresholds are **ordered**: $\alpha_1 < \alpha_2 < \cdots < \alpha_{J-1}$

::: {.callout-warning}
## The Sign Convention Varies by Software!

Some software (including `polr` and `clm`) uses the model:
$$\text{logit}(P(Y \leq j | \mathbf{X})) = \alpha_j - \boldsymbol{\beta}^T\mathbf{X}$$

Other software uses:
$$\text{logit}(P(Y \leq j | \mathbf{X})) = \alpha_j + \boldsymbol{\beta}^T\mathbf{X}$$

**This changes the interpretation of positive/negative coefficients!** See Section 5 for details.
:::

### Implied Category Probabilities

From the cumulative model, we can derive individual category probabilities:

$$P(Y = j | \mathbf{X}) = P(Y \leq j | \mathbf{X}) - P(Y \leq j-1 | \mathbf{X})$$

where $P(Y \leq 0) = 0$ and $P(Y \leq J) = 1$ by convention.


## 3. Coefficient Interpretation

### The Proportional Odds Interpretation

For `MASS::polr` and `ordinal::clm` (which use the **minus sign** convention):

| Coefficient Sign | Interpretation |
|------------------|----------------|
| $\beta > 0$ | Higher values of $X$ associated with **higher** outcome categories |
| $\beta < 0$ | Higher values of $X$ associated with **lower** outcome categories |
| $\exp(\beta)$ | **Cumulative odds ratio** for higher vs. lower categories |

### What is the Cumulative Odds Ratio?

The cumulative odds ratio $\exp(\beta)$ compares:

$$\text{OR} = \frac{\text{Odds}(Y > j | X = x+1)}{\text{Odds}(Y > j | X = x)} = \exp(\beta)$$

This is the same at **every cut point** (the proportional odds assumption).

### Interpretation Template

For a one-unit increase in $X$:

> "A one-unit increase in $X$ is associated with $\exp(\beta)$ times the odds of being in a **higher** category versus a **lower** category (or any categories above threshold $j$ versus at or below $j$)."

### Example Interpretations

| Scenario | $\beta$ | $\exp(\beta)$ | Interpretation |
|----------|---------|---------------|----------------|
| Treatment effect | 0.85 | 2.34 | Treatment associated with 2.34 times odds of higher category |
| Age effect | -0.03 | 0.97 | Each year of age associated with 3% lower odds of higher category |
| Dose effect | 1.20 | 3.32 | Each unit increase in dose associated with 3.32 times odds of higher category |

::: {.callout-note}
## Higher vs. Lower Categories

The cumulative odds ratio applies to the odds of being **above** any cutpoint versus **at or below** it. Because of the proportional odds assumption, this ratio is constant across all cutpoints.
:::


## 4. The Proportional Odds Assumption

### What It Means

The proportional odds assumption states that the effect of each predictor is **constant** across all cutpoints of the outcome. Mathematically:

$$\frac{\text{Odds}(Y > j | X = x+1)}{\text{Odds}(Y > j | X = x)} = \exp(\beta) \text{ for all } j$$

### Visual Representation

```{r}
#| echo: false
#| fig-height: 4
#| fig-width: 8

# Create illustration of proportional odds
x_vals <- seq(-3, 3, length.out = 100)
thresholds <- c(-1.5, 0, 1.5)
beta <- 0.8

# Calculate cumulative probabilities for X=0 and X=1
dat_plot <- expand.grid(x = x_vals, threshold = 1:3, X_group = c("X = 0", "X = 1"))
dat_plot$alpha <- thresholds[dat_plot$threshold]
dat_plot$X_val <- ifelse(dat_plot$X_group == "X = 0", 0, 1)
dat_plot$cumprob <- plogis(dat_plot$alpha - beta * dat_plot$X_val)
dat_plot$threshold_label <- paste("P(Y <=", dat_plot$threshold, ")")

ggplot(dat_plot[dat_plot$X_group %in% c("X = 0", "X = 1") & dat_plot$x == 0, ],
       aes(x = threshold, y = cumprob, color = X_group, group = X_group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  labs(x = "Cumulative Category (j)",
       y = "Cumulative Probability P(Y <= j)",
       color = "Predictor Value",
       title = "Proportional Odds: Parallel Cumulative Probability Curves") +
  scale_x_continuous(breaks = 1:3) +
  scale_y_continuous(limits = c(0, 1)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The curves are **parallel** on the logit scale, meaning the distance between them (the log-odds ratio) is constant.

### When Proportional Odds Fails

The assumption may be violated when:

- Different predictors have different effects at different severity levels
- There are "ceiling" or "floor" effects
- The categories are not truly ordinal
- Sample size is small at extreme categories

### Testing: The Brant Test

```{r}
#| eval: false

# Using the brant package
library(brant)
brant(polr_model)

# Interpretation:
# - Non-significant p-values suggest proportional odds is reasonable
# - Significant p-values indicate violation for that predictor
```

### Alternatives When Assumption Fails

| Alternative | Description |
|-------------|-------------|
| **Partial proportional odds** | Allow some predictors to have different effects at each cutpoint |
| **Multinomial logit** | Abandon ordering, treat as nominal |
| **Adjacent-category logit** | Model odds of adjacent categories |
| **Continuation-ratio logit** | Model sequential transitions |
| **Stereotype model** | Ordinal model without proportional odds |

The `ordinal::clm` package allows partial proportional odds models via the `nominal` argument.


## 5. Software Sign Conventions

::: {.callout-important}
## Critical: Know Your Software's Convention!

Different software packages use different sign conventions. Getting this wrong leads to completely reversed interpretations!
:::

### Sign Convention Comparison

| Package | Model Form | Positive $\beta$ Means | Notes |
|---------|------------|------------------------|-------|
| `MASS::polr` (R) | $\text{logit}(P(Y \leq j)) = \alpha_j - \beta X$ | Higher $X$ $\to$ **higher** $Y$ | Most common in R |
| `ordinal::clm` (R) | $\text{logit}(P(Y \leq j)) = \alpha_j - \beta X$ | Higher $X$ $\to$ **higher** $Y$ | Same as `polr` |
| `VGAM::vglm` (R) | $\text{logit}(P(Y \leq j)) = \alpha_j - \beta X$ | Higher $X$ $\to$ **higher** $Y$ | Uses "propodds" family |
| SAS PROC LOGISTIC | Depends on `DESCENDING` option | Check output! | Default may vary |
| Stata `ologit` | $\text{logit}(P(Y \leq j)) = \alpha_j - \beta X$ | Higher $X$ $\to$ **higher** $Y$ | Same convention |

### How to Verify Your Software's Convention

**Always verify with predicted probabilities:**

```{r}
# Create simple test data
set.seed(667)
test_dat <- data.frame(
  y = factor(c(rep(1, 20), rep(2, 30), rep(3, 30), rep(4, 20)),
             ordered = TRUE, levels = 1:4),
  x = c(rnorm(20, 0), rnorm(30, 1), rnorm(30, 2), rnorm(20, 3))
)

# Fit model
test_fit <- polr(y ~ x, data = test_dat)

# Get predictions for low and high x
pred_low <- predict(test_fit, newdata = data.frame(x = 0), type = "probs")
pred_high <- predict(test_fit, newdata = data.frame(x = 3), type = "probs")

cat("Predicted probabilities for x = 0:\n")
print(round(pred_low, 3))
cat("\nPredicted probabilities for x = 3:\n")
print(round(pred_high, 3))

# Check: If beta > 0 in polr, higher x should have more probability in higher categories
cat("\nCoefficient for x:", coef(test_fit), "\n")
cat("Beta > 0, so higher x should give more probability in higher categories\n")
cat("Sum of P(Y=3) + P(Y=4) for x=0:", sum(pred_low[3:4]), "\n")
cat("Sum of P(Y=3) + P(Y=4) for x=3:", sum(pred_high[3:4]), "\n")
```


## 6. Worked Example

### Pain Severity Data

Let's analyze a simulated clinical trial examining the effect of treatment on pain severity.

```{r}
set.seed(667)

# Simulate pain severity data
n <- 200

# Covariates
treatment <- factor(rep(c("Placebo", "Treatment"), each = n/2))
age <- round(rnorm(n, mean = 50, sd = 12))
baseline_pain <- sample(1:4, n, replace = TRUE, prob = c(0.1, 0.3, 0.4, 0.2))

# Generate ordinal outcome (1=none, 2=mild, 3=moderate, 4=severe)
# Treatment reduces pain (negative effect on severity)
# Higher baseline pain associated with higher post-treatment pain
eta <- -0.8 * (treatment == "Treatment") + 0.4 * (baseline_pain - 2) + 0.02 * (age - 50)

# Convert to ordinal using latent variable approach
latent <- eta + rlogis(n)
thresholds <- c(-2, 0, 2)
pain <- cut(latent, breaks = c(-Inf, thresholds, Inf), labels = c("None", "Mild", "Moderate", "Severe"))
pain <- ordered(pain, levels = c("None", "Mild", "Moderate", "Severe"))

# Create data frame
pain_data <- data.frame(
  pain = pain,
  treatment = treatment,
  age = age,
  baseline_pain = factor(baseline_pain, labels = c("None", "Mild", "Moderate", "Severe"))
)

# View data summary
table(pain_data$pain, pain_data$treatment)
```

### Fit the Proportional Odds Model

```{r}
# Fit with MASS::polr
fit_polr <- polr(pain ~ treatment + age + baseline_pain, data = pain_data, Hess = TRUE)

# Summary
summary(fit_polr)
```

### Extract and Interpret Coefficients

```{r}
# Extract coefficients and standard errors
coef_table <- coef(summary(fit_polr))
print(coef_table)

# Calculate odds ratios and 95% CIs
beta <- coef_table[, "Value"]
se <- coef_table[, "Std. Error"]

# Create results table
results <- data.frame(
  Variable = names(beta),
  Coefficient = beta,
  SE = se,
  OR = exp(beta),
  OR_Lower = exp(beta - 1.96 * se),
  OR_Upper = exp(beta + 1.96 * se),
  z_value = beta / se,
  p_value = 2 * (1 - pnorm(abs(beta / se)))
)

print(results, digits = 3)
```

### Interpretation

```{r}
#| echo: false

trt_coef <- coef(fit_polr)["treatmentTreatment"]
trt_or <- exp(trt_coef)
```

**Treatment effect:**

- Coefficient = `r round(trt_coef, 3)`
- Odds Ratio = `r round(trt_or, 3)`

Since we are using `polr` (minus sign convention) and $\beta < 0$:

> "Patients receiving treatment have `r round(trt_or, 2)` times the odds of being in a **higher** (more severe) pain category compared to placebo. Equivalently, treatment is associated with `r round(1/trt_or, 2)` times the odds of being in a **lower** (less severe) pain category."

Or more intuitively:

> "Treatment reduces the odds of higher pain severity by `r round((1-trt_or)*100)`% compared to placebo."


## 7. Predicted Probabilities

### Calculating Category Probabilities

```{r}
# Get predicted probabilities for all observations
pred_probs <- predict(fit_polr, type = "probs")
head(pred_probs)
```

### Predictions for Specific Covariate Values

```{r}
# Create prediction scenarios
new_data <- expand.grid(
  treatment = c("Placebo", "Treatment"),
  age = 50,
  baseline_pain = "Moderate"
)

# Get predicted probabilities
pred_scenarios <- predict(fit_polr, newdata = new_data, type = "probs")
pred_scenarios <- as.data.frame(pred_scenarios)
pred_scenarios <- cbind(new_data[, "treatment", drop = FALSE], pred_scenarios)

print(pred_scenarios, digits = 3)
```

### Visualization of Predicted Probabilities

```{r}
#| fig-height: 5
#| fig-width: 8

# Create data for plotting
plot_data <- expand.grid(
  treatment = c("Placebo", "Treatment"),
  age = 50,
  baseline_pain = c("None", "Mild", "Moderate", "Severe")
)

pred_plot <- predict(fit_polr, newdata = plot_data, type = "probs")
plot_data <- cbind(plot_data, pred_plot)

# Reshape for plotting
library(tidyr)
plot_long <- pivot_longer(plot_data, cols = c("None", "Mild", "Moderate", "Severe"),
                          names_to = "Pain_Level", values_to = "Probability")
plot_long$Pain_Level <- factor(plot_long$Pain_Level,
                                levels = c("None", "Mild", "Moderate", "Severe"))

ggplot(plot_long, aes(x = baseline_pain, y = Probability, fill = Pain_Level)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~ treatment) +
  scale_fill_brewer(palette = "YlOrRd", name = "Post-Treatment\nPain Level") +
  labs(x = "Baseline Pain Level", y = "Predicted Probability",
       title = "Predicted Pain Outcomes by Treatment and Baseline Pain",
       subtitle = "Age = 50 years") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Cumulative Probability Curves

```{r}
#| fig-height: 4
#| fig-width: 8

# Calculate cumulative probabilities across age range
age_range <- data.frame(
  treatment = rep(c("Placebo", "Treatment"), each = 50),
  age = rep(seq(30, 70, length.out = 50), 2),
  baseline_pain = "Moderate"
)

# Get cumulative probabilities
pred_age <- predict(fit_polr, newdata = age_range, type = "probs")
age_range <- cbind(age_range, pred_age)

# Calculate cumulative probabilities
age_range$cum_mild <- age_range$None + age_range$Mild
age_range$cum_moderate <- age_range$cum_mild + age_range$Moderate

# Plot
plot_cum <- pivot_longer(age_range,
                         cols = c("None", "cum_mild", "cum_moderate"),
                         names_to = "Cumulative", values_to = "Prob")
plot_cum$Cumulative <- factor(plot_cum$Cumulative,
                              levels = c("None", "cum_mild", "cum_moderate"),
                              labels = c("P(Y <= None)", "P(Y <= Mild)", "P(Y <= Moderate)"))

ggplot(plot_cum, aes(x = age, y = Prob, color = Cumulative, linetype = treatment)) +
  geom_line(linewidth = 1) +
  labs(x = "Age (years)", y = "Cumulative Probability",
       title = "Cumulative Probability Curves by Age and Treatment",
       subtitle = "Baseline pain = Moderate",
       color = "Cumulative Probability",
       linetype = "Treatment") +
  theme_minimal() +
  theme(legend.position = "bottom")
```


## 8. Using emmeans for Adjusted Probabilities

The `emmeans` package provides an elegant way to obtain adjusted predicted probabilities and contrasts.

### Adjusted Probabilities by Treatment

```{r}
# Get adjusted predicted probabilities
em_probs <- emmeans(fit_polr, ~ pain | treatment, mode = "prob")
summary(em_probs)
```

### Pairwise Comparisons on Odds Ratio Scale

```{r}
# Treatment comparison on odds ratio scale
em_or <- emmeans(fit_polr, ~ treatment, mode = "latent")
pairs(em_or, type = "response")
```

### Marginal Effects

```{r}
# Joint reference grid
em_joint <- emmeans(fit_polr, ~ treatment, mode = "latent")
summary(em_joint)

# Contrast (log-odds scale)
contrast(em_joint, method = "pairwise")
```

### Plotting emmeans Results

```{r}
#| fig-height: 4
#| fig-width: 7

# Extract probability estimates for plotting
em_summary <- as.data.frame(summary(em_probs))

ggplot(em_summary, aes(x = pain, y = prob, fill = treatment)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL),
                position = position_dodge(0.7), width = 0.25) +
  labs(x = "Pain Severity", y = "Adjusted Probability (with 95% CI)",
       title = "Adjusted Predicted Probabilities by Treatment",
       fill = "Treatment Group") +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal()
```


## 9. Common Mistakes

### Mistake 1: Getting the Sign Wrong

```{r}
#| eval: false

# WRONG: Interpreting positive coefficient as lower outcome
# "Treatment (beta = 0.8) is associated with lower pain"

# CORRECT for polr/clm: Positive coefficient means HIGHER category
# "Treatment (beta = 0.8) is associated with HIGHER pain severity"
```

**Always verify with predicted probabilities!**

### Mistake 2: Interpreting as Probability

```{r}
#| eval: false

# WRONG
# "A one-unit increase in X increases the probability of higher category by exp(beta)"

# CORRECT
# "A one-unit increase in X multiplies the ODDS of higher category by exp(beta)"
```

### Mistake 3: Ignoring Proportional Odds Violation

```{r}
#| eval: false

# WRONG: Blindly reporting results without checking
summary(fit_polr)

# CORRECT: Test the assumption
library(brant)
brant(fit_polr)  # Check for violations

# If violated, consider alternatives:
# - Partial proportional odds (clm with nominal terms)
# - Multinomial regression
# - Report with caveat about violation
```

### Mistake 4: Forgetting Software Conventions

```{r}
# Different packages may give different signs!
# Always document which package and verify interpretation

# Example: Compare polr and clm (should be same)
fit_clm <- clm(pain ~ treatment + age + baseline_pain, data = pain_data)

cat("polr coefficient for treatment:", coef(fit_polr)["treatmentTreatment"], "\n")
cat("clm coefficient for treatment:", coef(fit_clm)["treatmentTreatment"], "\n")
```

### Mistake 5: Treating Ordinal as Continuous

```{r}
#| eval: false

# WRONG: Linear regression on ordinal outcome
lm(as.numeric(pain) ~ treatment, data = pain_data)

# This assumes equal intervals between categories!

# CORRECT: Use ordinal regression
polr(pain ~ treatment, data = pain_data)
```

### Summary of Common Mistakes

| Mistake | Problem | Solution |
|---------|---------|----------|
| Sign reversal | Wrong interpretation direction | Verify with predicted probabilities |
| Probability vs odds | Misinterpret coefficient magnitude | Report as odds ratio |
| Ignoring proportional odds | Invalid inference | Test with Brant test |
| Wrong software convention | Completely wrong interpretation | Check package documentation |
| Treating as continuous | Assumes equal intervals | Use proper ordinal model |


## 10. Quick Reference Card

### Interpretation Template

For `MASS::polr` and `ordinal::clm`:

::: {.callout-tip}
## Standard Interpretation

"A one-unit increase in $X$ is associated with $\exp(\beta)$ = **[value]** times the odds of being in a **higher** versus **lower** category of $Y$."

Or equivalently:

"A one-unit increase in $X$ is associated with $\exp(\beta)$ times the odds of being **above** any given threshold versus **at or below** that threshold."
:::

### Quick Formulas

| Quantity | Formula |
|----------|---------|
| Odds ratio | $\text{OR} = \exp(\beta)$ |
| 95% CI for OR | $\exp(\beta \pm 1.96 \times \text{SE})$ |
| Probability $P(Y \leq j)$ | $\text{logit}^{-1}(\alpha_j - \beta X)$ |
| Probability $P(Y = j)$ | $P(Y \leq j) - P(Y \leq j-1)$ |

### Essential R Code

```{r}
#| eval: false

# Fit model
library(MASS)
fit <- polr(outcome ~ predictor, data = dat, Hess = TRUE)

# Summary with p-values
coef_table <- coef(summary(fit))
p_values <- 2 * (1 - pnorm(abs(coef_table[, "t value"])))

# Odds ratios with CIs
exp(coef(fit))                              # OR
exp(confint(fit))                           # 95% CI for OR

# Predicted probabilities
predict(fit, type = "probs")                # Category probs
predict(fit, type = "class")                # Predicted class

# Test proportional odds
library(brant)
brant(fit)

# Adjusted predictions with emmeans
library(emmeans)
emmeans(fit, ~ predictor, mode = "prob")
```

### Decision Guide

```
Is outcome ordered?
  --> No  --> Use multinomial logistic regression
  --> Yes --> Continue

Are intervals between categories equal?
  --> Yes --> Can consider linear regression (with caution)
  --> No  --> Use ordinal regression

Does proportional odds hold?
  --> Yes --> Standard cumulative logit model (polr, clm)
  --> No  --> Partial proportional odds, multinomial, or alternative models
```

### Package Quick Reference

```{r}
#| eval: false

# Core ordinal regression
library(MASS)        # polr() - standard cumulative logit
library(ordinal)     # clm() - more flexible, allows partial proportional odds

# Diagnostics
library(brant)       # brant() - test proportional odds assumption

# Marginal means and contrasts
library(emmeans)     # emmeans() - adjusted predictions and comparisons

# Visualization
library(ggplot2)     # Plotting
library(effects)     # effect() - effect plots for ordinal models
```

::: {.callout-tip}
## Final Tips

1. **Always verify sign convention** with predicted probabilities before interpreting
2. **Test proportional odds assumption** before trusting a single coefficient
3. **Report odds ratios** with confidence intervals, not just coefficients
4. **Visualize predicted probabilities** to communicate results effectively
5. **Use emmeans** for clean marginal predictions and treatment comparisons
:::
