---
title: "Random Slopes: When to Add and How to Interpret"
subtitle: "Beyond Random Intercepts in Mixed Models"
format: html
toc: true
embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(nlme)
library(lme4)
library(ggplot2)
library(dplyr)
library(gridExtra)

set.seed(667)
```

## 1. Random Intercepts vs Random Slopes

### Conceptual Distinction

**Random Intercepts Only (RI):**

- All subjects share the *same* slope (rate of change over time)
- Subjects differ only in their starting points (baseline values)
- The trajectories are parallel lines shifted vertically

**Random Intercepts + Random Slopes (RI + RS):**

- Subjects have *different* slopes AND different starting points
- Each subject has their own trajectory
- Lines can cross each other---not parallel

### Visual Comparison

```{r}
#| label: fig-ri-vs-rs
#| fig-cap: "Comparison of Random Intercept Only vs Random Intercept + Slope Models"
#| fig-width: 10
#| fig-height: 5

# Simulate data for visualization
n_subj <- 8
n_time <- 5
time_points <- 0:4

# Random intercepts only (parallel lines)
ri_data <- data.frame(
  id = rep(1:n_subj, each = n_time),
  time = rep(time_points, n_subj)
)
ri_intercepts <- rnorm(n_subj, mean = 10, sd = 3)
ri_data$y <- ri_intercepts[ri_data$id] + 2 * ri_data$time + rnorm(n_subj * n_time, 0, 0.5)

# Random intercepts + slopes (non-parallel lines)
rs_data <- data.frame(
  id = rep(1:n_subj, each = n_time),
  time = rep(time_points, n_subj)
)
rs_intercepts <- rnorm(n_subj, mean = 10, sd = 3)
rs_slopes <- rnorm(n_subj, mean = 2, sd = 1.5)  # Slopes vary!
rs_data$y <- rs_intercepts[rs_data$id] + rs_slopes[rs_data$id] * rs_data$time +
             rnorm(n_subj * n_time, 0, 0.5)

# Create plots
p1 <- ggplot(ri_data, aes(x = time, y = y, group = factor(id), color = factor(id))) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(title = "Random Intercepts Only",
       subtitle = "Parallel trajectories (same slope for all)",
       x = "Time", y = "Outcome") +
  theme_minimal() +
  theme(legend.position = "none")

p2 <- ggplot(rs_data, aes(x = time, y = y, group = factor(id), color = factor(id))) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(title = "Random Intercepts + Slopes",
       subtitle = "Non-parallel trajectories (slopes vary)",
       x = "Time", y = "Outcome") +
  theme_minimal() +
  theme(legend.position = "none")

grid.arrange(p1, p2, ncol = 2)
```

**Key Insight:** In the left panel, subjects start at different places but change at the same rate. In the right panel, subjects both start differently AND change at different rates---some increase quickly, others slowly, and some may even decrease.

***

## 2. When to Add Random Slopes

Use the following diagnostic signals to decide whether random slopes are needed:

| Signal | What it Means | Action |
|--------|---------------|--------|
| **Fanning pattern** in residuals vs time | Subject-level heterogeneity in slopes | Add random slope for time |
| **Theoretical expectation** of varying trajectories | Domain knowledge suggests individual variation | Include random slope a priori |
| **Significant LRT** comparing RI vs RI+RS | Random slope improves model fit | Retain the random slope |
| **Random slope variance** significantly > 0 | Meaningful between-subject variability in slopes | Keep the random slope |
| **Spaghetti plot** shows crossing/diverging lines | Visual evidence of slope heterogeneity | Strong candidate for random slope |

### Decision Framework

1. **Plot individual trajectories** (spaghetti plot)
   - Do lines cross or fan out? Consider random slopes

2. **Fit RI-only model**, examine residuals vs time
   - Fanning pattern present? Add random slopes

3. **Compare RI vs RI+RS models** via LRT (use ML, not REML)
   - Significant improvement? Keep random slopes

4. **Check if model converges** with random slopes
   - Convergence issues? May need simpler structure

***

## 3. The Model

### Mathematical Specification

The linear mixed model with random intercepts and slopes is:

$$
Y_{ij} = (\beta_0 + b_{0i}) + (\beta_1 + b_{1i}) \cdot \text{time}_{ij} + \varepsilon_{ij}
$$

Where:

- $Y_{ij}$ = outcome for subject $i$ at time $j$
- $\beta_0$ = population-average intercept (mean at time = 0)
- $\beta_1$ = population-average slope (mean rate of change)
- $b_{0i}$ = subject-specific deviation from population intercept
- $b_{1i}$ = subject-specific deviation from population slope
- $\varepsilon_{ij}$ = residual error

### Random Effects Distribution

The random effects follow a multivariate normal distribution:

$$
\begin{pmatrix} b_{0i} \\ b_{1i} \end{pmatrix} \sim \text{MVN}\left( \begin{pmatrix} 0 \\ 0 \end{pmatrix}, \mathbf{D} \right)
$$

Where the variance-covariance matrix is:

$$
\mathbf{D} = \begin{pmatrix} \sigma^2_{b_0} & \sigma_{b_{01}} \\ \sigma_{b_{01}} & \sigma^2_{b_1} \end{pmatrix}
$$

### Interpreting the Correlation

The correlation between random intercepts and slopes:

$$
\rho_{01} = \frac{\sigma_{b_{01}}}{\sigma_{b_0} \cdot \sigma_{b_1}}
$$

| Correlation | Interpretation |
|-------------|----------------|
| $\rho_{01} > 0$ (positive) | Subjects who start higher tend to increase faster |
| $\rho_{01} < 0$ (negative) | Subjects who start higher tend to increase slower (regression to mean) |
| $\rho_{01} \approx 0$ | Starting point and rate of change are independent |

**Example:** In growth studies, negative correlation often occurs---children who are taller at baseline tend to grow more slowly (they're closer to their adult height).

***

## 4. Variance Components Interpretation

| Parameter | Symbol | Interpretation |
|-----------|--------|----------------|
| Random intercept variance | $\sigma^2_{b_0}$ | Between-subject variance **at time = 0**; how much subjects differ at baseline |
| Random slope variance | $\sigma^2_{b_1}$ | Between-subject variance **in rates of change**; how much slopes vary across subjects |
| Covariance | $\sigma_{b_{01}}$ | Association between starting point and rate of change |
| Correlation | $\rho_{01}$ | Standardized covariance (bounded -1 to 1) |
| Residual variance | $\sigma^2$ | Within-subject variance not explained by random effects |

### Important Note on Centering

**The interpretation of $\sigma^2_{b_0}$ depends on how time is coded:**

- If `time` starts at 0: $\sigma^2_{b_0}$ = variance at baseline
- If `time` is centered at the mean: $\sigma^2_{b_0}$ = variance at the mean time point
- This also affects the intercept-slope covariance!

**Recommendation:** Always think carefully about centering time to make variance components interpretable.

***

## 5. R Code Syntax

### Package Comparison

**Random intercept only:**

- nlme: `random = ~1|id`
- lme4: `(1|id)`

**Random intercept + slope:**

- nlme: `random = ~time|id`
- lme4: `(1 + time|id)`

**Random slope only** (rare):

- nlme: `random = ~0 + time|id`
- lme4: `(0 + time|id)`

**Uncorrelated RI + RS:**

- nlme: `random = list(id = pdDiag(~time))`
- lme4: `(1|id) + (0 + time|id)`

### nlme Examples

```{r}
#| eval: false

# Random intercept only
fit_ri <- lme(y ~ time + treatment,
              random = ~1|id,
              data = mydata)

# Random intercept + slope (correlated)
fit_rs <- lme(y ~ time + treatment,
              random = ~time|id,
              data = mydata)

# Random intercept + slope (uncorrelated - diagonal D matrix)
fit_rs_uncor <- lme(y ~ time + treatment,
                    random = list(id = pdDiag(~time)),
                    data = mydata)
```

### lme4 Examples

```{r}
#| eval: false

# Random intercept only
fit_ri <- lmer(y ~ time + treatment + (1|id), data = mydata)

# Random intercept + slope (correlated)
fit_rs <- lmer(y ~ time + treatment + (1 + time|id), data = mydata)

# Random intercept + slope (uncorrelated)
fit_rs_uncor <- lmer(y ~ time + treatment + (1|id) + (0 + time|id), data = mydata)
```

### Extracting Variance Components

```{r}
#| eval: false

# nlme
VarCorr(fit_rs)  # Shows variances and correlations
intervals(fit_rs)  # Confidence intervals

# lme4
VarCorr(fit_rs)  # Shows variances and correlations
as.data.frame(VarCorr(fit_rs))  # As a data frame
confint(fit_rs, oldNames = FALSE)  # Profile likelihood CIs
```

***

## 6. Diagnostic Check: The Fanning Pattern

The "fanning pattern" is a key diagnostic for detecting the need for random slopes.

### What to Look For

```{r}
#| label: fig-fanning
#| fig-cap: "Detecting the Fanning Pattern in Residuals"
#| fig-width: 10
#| fig-height: 5

# Simulate data with heterogeneous slopes
n_subj <- 50
n_time <- 6

sim_data <- data.frame(
  id = rep(1:n_subj, each = n_time),
  time = rep(0:(n_time-1), n_subj)
)

# True model has random slopes
true_int <- rnorm(n_subj, 50, 5)
true_slope <- rnorm(n_subj, 3, 2)  # Substantial slope variability
sim_data$y <- true_int[sim_data$id] + true_slope[sim_data$id] * sim_data$time +
              rnorm(n_subj * n_time, 0, 2)

# Fit RI-only model (misspecified)
fit_ri_wrong <- lme(y ~ time, random = ~1|id, data = sim_data)

# Fit correct RI+RS model
fit_rs_correct <- lme(y ~ time, random = ~time|id, data = sim_data)

# Get residuals
sim_data$resid_ri <- residuals(fit_ri_wrong)
sim_data$resid_rs <- residuals(fit_rs_correct)

# Plot residuals
p1 <- ggplot(sim_data, aes(x = time, y = resid_ri)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = FALSE, color = "red") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "RI Only: Fanning Pattern Present",
       subtitle = "Residual variance increases with time",
       x = "Time", y = "Residuals") +
  theme_minimal()

p2 <- ggplot(sim_data, aes(x = time, y = resid_rs)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = FALSE, color = "blue") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "RI + RS: Fanning Resolved",
       subtitle = "Residual variance constant across time",
       x = "Time", y = "Residuals") +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```

### Interpreting the Pattern

**Left panel (RI only):** The spread of residuals increases (or decreases) with time. This "fanning" or "funnel" shape indicates that the model is not capturing subject-specific variation in slopes.

**Right panel (RI + RS):** After adding random slopes, the residuals show constant variance across time. The fanning disappears because subject-specific slopes are now modeled.

***

## 7. Boundary Tests for Variance Components

### The Problem

When testing whether a random slope is needed:

$$
H_0: \sigma^2_{b_1} = 0 \quad \text{vs} \quad H_1: \sigma^2_{b_1} > 0
$$

The null hypothesis places the parameter on the boundary of the parameter space (variance cannot be negative). This violates standard LRT assumptions.

### The Solution: Mixture Chi-Square Distribution

Under the null, the LRT statistic follows a **mixture of chi-square distributions**, not a standard chi-square.

For testing one variance component:

$$
\text{LRT} \sim 0.5 \cdot \chi^2_0 + 0.5 \cdot \chi^2_1
$$

This means the p-value from a standard `anova()` is **conservative** (too large).

### Practical Approaches

**Option 1: Divide the p-value by 2**

```{r}
#| eval: false

# Compare RI vs RI+RS models (must use ML, not REML!)
fit_ri_ml <- lme(y ~ time, random = ~1|id, data = mydata, method = "ML")
fit_rs_ml <- lme(y ~ time, random = ~time|id, data = mydata, method = "ML")

lrt_result <- anova(fit_ri_ml, fit_rs_ml)
corrected_pvalue <- lrt_result$`p-value`[2] / 2
```

**Option 2: Use the exact mixture distribution**

For adding both a variance and covariance (2 parameters):

$$
\text{LRT} \sim 0.5 \cdot \chi^2_1 + 0.5 \cdot \chi^2_2
$$

```{r}
#| eval: false

# Critical value for mixture (alpha = 0.05)
# P(0.5*chi2_1 + 0.5*chi2_2 > cv) = 0.05
# Approximately cv = 5.14 for 2 df case

lrt_stat <- -2 * (logLik(fit_ri_ml) - logLik(fit_rs_ml))
pval_mixture <- 0.5 * pchisq(lrt_stat, df = 1, lower.tail = FALSE) +
                0.5 * pchisq(lrt_stat, df = 2, lower.tail = FALSE)
```

### Important Reminder

**Always use ML (not REML) when comparing models with different fixed OR random effects structures via LRT.**

```{r}
#| eval: false

# WRONG: Using REML for model comparison
fit1 <- lme(y ~ time, random = ~1|id, data = mydata)  # REML is default
fit2 <- lme(y ~ time, random = ~time|id, data = mydata)
anova(fit1, fit2)  # Invalid comparison!

# CORRECT: Use ML
fit1_ml <- lme(y ~ time, random = ~1|id, data = mydata, method = "ML")
fit2_ml <- lme(y ~ time, random = ~time|id, data = mydata, method = "ML")
anova(fit1_ml, fit2_ml)  # Valid comparison
```

***

## 8. BLUPs for Subject-Specific Predictions

### What are BLUPs?

**BLUP** = Best Linear Unbiased Predictor

BLUPs provide estimates of the subject-specific random effects ($b_{0i}$, $b_{1i}$). They are "predictions" rather than "estimates" because random effects are random variables, not fixed parameters.

### Key Properties

1. **Shrinkage:** BLUPs are "shrunk" toward zero (the population mean of random effects)
2. **Reliability-weighted:** Subjects with more observations or less noise get BLUPs closer to their empirical values
3. **Borrowing strength:** Information from all subjects informs each individual's BLUP

### Extracting BLUPs in R

```{r}
#| eval: false

# nlme
ranef(fit_rs)           # Random effects (deviations from population)
coef(fit_rs)            # Subject-specific coefficients (β + b)

# lme4
ranef(fit_rs)           # Random effects
coef(fit_rs)$id         # Subject-specific coefficients
```

### Predictions: Population vs Subject-Specific

```{r}
#| eval: false

# nlme predictions
predict(fit_rs, level = 0)  # Population-average (fixed effects only)
predict(fit_rs, level = 1)  # Subject-specific (fixed + random effects)

# lme4 predictions
predict(fit_rs, re.form = NA)   # Population-average
predict(fit_rs, re.form = NULL) # Subject-specific (default)
```

| Level | What it Gives | Use Case |
|-------|---------------|----------|
| `level = 0` / `re.form = NA` | $\hat{Y} = \hat{\beta}_0 + \hat{\beta}_1 \cdot \text{time}$ | Population-average trajectory |
| `level = 1` / `re.form = NULL` | $\hat{Y} = (\hat{\beta}_0 + \hat{b}_{0i}) + (\hat{\beta}_1 + \hat{b}_{1i}) \cdot \text{time}$ | Individual predictions |

### Visualizing Shrinkage

```{r}
#| label: fig-shrinkage
#| fig-cap: "BLUP Shrinkage: Subject-Specific Estimates Pulled Toward Population Mean"
#| fig-width: 8
#| fig-height: 6

# Using the earlier simulated data
blups <- ranef(fit_rs_correct)

# Calculate empirical (OLS) slopes for each subject
empirical_slopes <- sim_data %>%
  group_by(id) %>%
  summarize(
    ols_slope = coef(lm(y ~ time))[2],
    n_obs = n()
  )

# Compare to BLUPs
comparison <- data.frame(
  id = 1:n_subj,
  ols_slope = empirical_slopes$ols_slope,
  blup_slope = fixef(fit_rs_correct)["time"] + blups$time
)

# Population mean slope
pop_slope <- fixef(fit_rs_correct)["time"]

ggplot(comparison, aes(x = ols_slope, y = blup_slope)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
  geom_hline(yintercept = pop_slope, color = "red", linetype = "dotted") +
  geom_vline(xintercept = pop_slope, color = "red", linetype = "dotted") +
  labs(title = "Shrinkage of Subject-Specific Slopes",
       subtitle = "BLUPs are pulled toward the population mean (red lines)",
       x = "Empirical (OLS) Slope",
       y = "BLUP Slope") +
  theme_minimal() +
  annotate("text", x = max(comparison$ols_slope) - 1, y = pop_slope + 0.3,
           label = "Population mean", color = "red", size = 3)
```

**Interpretation:** Points lie between the 45-degree line (no shrinkage) and the population mean (complete shrinkage). Extreme OLS estimates are pulled more strongly toward the mean.

***
## 9. Worked Example

### The Data: Longitudinal Growth Study

We'll simulate a growth study where children are measured at multiple time points. We expect individual variation in both baseline height and growth rates.

```{r}
#| label: example-data

set.seed(667)

# Simulation parameters
n_children <- 40
n_visits <- 5
ages <- c(6, 7, 8, 9, 10)  # Ages 6-10

# Generate data
growth_data <- data.frame(
  child_id = rep(1:n_children, each = n_visits),
  age = rep(ages, n_children)
)

# True parameters
beta0 <- 110     # Mean height at age 6
beta1 <- 5       # Mean growth rate (cm/year)
sigma_b0 <- 6    # SD of random intercepts
sigma_b1 <- 1.2  # SD of random slopes
rho <- -0.4      # Correlation (taller kids grow slower)
sigma_e <- 1.5   # Residual SD

# Generate correlated random effects
library(MASS)
D <- matrix(c(sigma_b0^2, rho*sigma_b0*sigma_b1,
              rho*sigma_b0*sigma_b1, sigma_b1^2), nrow = 2)
random_effects <- mvrnorm(n_children, mu = c(0, 0), Sigma = D)

# Center age at 6 (baseline)
growth_data$age_c <- growth_data$age - 6

# Generate outcomes
growth_data$height <- beta0 + random_effects[growth_data$child_id, 1] +
                      (beta1 + random_effects[growth_data$child_id, 2]) * growth_data$age_c +
                      rnorm(n_children * n_visits, 0, sigma_e)

head(growth_data, 10)
```

### Step 1: Visualize the Data

```{r}
#| label: fig-spaghetti
#| fig-cap: "Individual Growth Trajectories"
#| fig-width: 8
#| fig-height: 6

ggplot(growth_data, aes(x = age, y = height, group = child_id)) +
  geom_line(alpha = 0.5, color = "steelblue") +
  geom_point(alpha = 0.5, size = 1) +
  labs(title = "Spaghetti Plot: Individual Growth Curves",
       subtitle = "Lines cross, suggesting varying slopes",
       x = "Age (years)", y = "Height (cm)") +
  theme_minimal()
```

**Observation:** The trajectories are not parallel---lines cross each other, suggesting children grow at different rates.

### Step 2: Fit Random Intercept Only Model

```{r}
#| label: fit-ri

# Random intercept only
fit_ri <- lme(height ~ age_c,
              random = ~1|child_id,
              data = growth_data)

summary(fit_ri)
```

### Step 3: Check for Fanning Pattern

```{r}
#| label: fig-ri-diagnostics
#| fig-cap: "Residual Diagnostics for RI Model"
#| fig-width: 8
#| fig-height: 5

growth_data$resid_ri <- residuals(fit_ri)

ggplot(growth_data, aes(x = age_c, y = resid_ri)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = TRUE, color = "red") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Time: Random Intercept Model",
       subtitle = "Look for fanning pattern",
       x = "Age (centered)", y = "Residuals") +
  theme_minimal()
```

### Step 4: Fit Random Intercept + Slope Model

```{r}
#| label: fit-rs

# Random intercept + slope
fit_rs <- lme(height ~ age_c,
              random = ~age_c|child_id,
              data = growth_data)

summary(fit_rs)
```

### Step 5: Compare Variance Components

```{r}
#| label: variance-comparison

# Extract variance components
cat("=== Random Intercept Only ===\n")
VarCorr(fit_ri)

cat("\n=== Random Intercept + Slope ===\n")
VarCorr(fit_rs)
```

**Interpretation of RI+RS variance components:**

- **Random intercept SD:** Variation in height at baseline (age 6)
- **Random slope SD:** Variation in growth rates across children
- **Correlation:** Negative correlation suggests taller children at baseline grow more slowly

### Step 6: Model Comparison via LRT

```{r}
#| label: model-comparison

# Must use ML for valid LRT
fit_ri_ml <- lme(height ~ age_c, random = ~1|child_id,
                 data = growth_data, method = "ML")
fit_rs_ml <- lme(height ~ age_c, random = ~age_c|child_id,
                 data = growth_data, method = "ML")

# LRT
lrt <- anova(fit_ri_ml, fit_rs_ml)
print(lrt)

# Corrected p-value for boundary test (conservative)
cat("\nCorrected p-value (boundary adjustment):",
    lrt$`p-value`[2] / 2, "\n")
```

### Step 7: Check Residuals After Adding Random Slopes

```{r}
#| label: fig-rs-diagnostics
#| fig-cap: "Residual Comparison: Before and After Adding Random Slopes"
#| fig-width: 10
#| fig-height: 5

growth_data$resid_rs <- residuals(fit_rs)

p1 <- ggplot(growth_data, aes(x = age_c, y = resid_ri)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = FALSE, color = "red") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  ylim(c(-5, 5)) +
  labs(title = "RI Only", x = "Age (centered)", y = "Residuals") +
  theme_minimal()

p2 <- ggplot(growth_data, aes(x = age_c, y = resid_rs)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = FALSE, color = "blue") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  ylim(c(-5, 5)) +
  labs(title = "RI + RS", x = "Age (centered)", y = "Residuals") +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```

### Step 8: Extract and Plot BLUPs

```{r}
#| label: fig-blups
#| fig-cap: "Subject-Specific Predictions from BLUPs"
#| fig-width: 10
#| fig-height: 6

# Get BLUPs
blups <- ranef(fit_rs)
head(blups)

# Create prediction data
pred_data <- expand.grid(
  child_id = unique(growth_data$child_id),
  age_c = seq(0, 4, by = 0.5)
)
pred_data$age <- pred_data$age_c + 6

# Population average prediction
pred_data$pop_pred <- predict(fit_rs, newdata = pred_data, level = 0)

# Subject-specific predictions
pred_data$subj_pred <- predict(fit_rs, newdata = pred_data, level = 1)

# Plot
ggplot() +
  # Subject-specific predictions
  geom_line(data = pred_data,
            aes(x = age, y = subj_pred, group = child_id),
            alpha = 0.3, color = "steelblue") +
  # Population average
  geom_line(data = pred_data[pred_data$child_id == 1, ],
            aes(x = age, y = pop_pred),
            color = "red", linewidth = 2) +
  # Observed data
  geom_point(data = growth_data,
             aes(x = age, y = height),
             alpha = 0.3, size = 1) +
  labs(title = "Subject-Specific Predictions vs Population Average",
       subtitle = "Blue = individual BLUPs, Red = population average",
       x = "Age (years)", y = "Height (cm)") +
  theme_minimal()
```

### Step 9: Examine Intercept-Slope Correlation

```{r}
#| label: fig-correlation
#| fig-cap: "Correlation Between Random Intercepts and Slopes"
#| fig-width: 7
#| fig-height: 6

# Extract subject-specific coefficients
subj_coef <- coef(fit_rs)

ggplot(subj_coef, aes(x = `(Intercept)`, y = age_c)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +
  labs(title = "Intercept-Slope Correlation",
       subtitle = paste("Correlation =", round(cor(subj_coef$`(Intercept)`, subj_coef$age_c), 3)),
       x = "Subject-Specific Intercept (Height at Age 6)",
       y = "Subject-Specific Slope (Growth Rate)") +
  theme_minimal()
```

**Interpretation:** The negative correlation indicates that children who are taller at age 6 tend to grow more slowly---consistent with the idea that they are closer to their adult height.

***

## 10. Common Mistakes

### Mistake 1: Adding Too Many Random Effects

**Problem:** Including random slopes for every predictor leads to convergence issues.

```{r}
#| eval: false

# Problematic: Too complex random effects structure
fit_overfit <- lmer(y ~ time * treatment + (1 + time + treatment + time:treatment | id),
                    data = mydata)
# Often fails to converge or produces singular fits
```

**Solution:** Start simple, add complexity only when justified by data and theory.

### Mistake 2: Not Centering Time

**Problem:** When time doesn't start at 0, the random intercept variance may not be interpretable.

```{r}
#| label: centering-example

# Example: time measured as years 2020, 2021, 2022...
# Random intercept variance would be at "year 0" (meaningless!)

# Better: Center time at baseline
# growth_data$time_c <- growth_data$year - min(growth_data$year)
```

**Solution:** Center time at a meaningful reference point (usually baseline).

### Mistake 3: Ignoring the Intercept-Slope Correlation

**Problem:** Forcing uncorrelated random effects when correlation exists.

```{r}
#| eval: false

# Forces zero correlation (may be inappropriate)
fit_uncor <- lmer(y ~ time + (1|id) + (0 + time|id), data = mydata)

# Allows correlation (usually preferred unless theory suggests otherwise)
fit_cor <- lmer(y ~ time + (1 + time|id), data = mydata)
```

**Solution:** Generally allow correlation unless you have strong theoretical reasons not to.

### Mistake 4: Using LRT with REML

**Problem:** REML likelihoods are not comparable across models with different fixed effects.

```{r}
#| eval: false

# WRONG
fit1 <- lme(y ~ time, random = ~1|id, data = mydata)  # REML default
fit2 <- lme(y ~ time, random = ~time|id, data = mydata)
anova(fit1, fit2)  # Invalid!

# CORRECT
fit1_ml <- lme(y ~ time, random = ~1|id, data = mydata, method = "ML")
fit2_ml <- lme(y ~ time, random = ~time|id, data = mydata, method = "ML")
anova(fit1_ml, fit2_ml)  # Valid
```

**Rule:** Use ML for model comparison, REML for final parameter estimation.

***

## The "Keep It Maximal" vs "Parsimonious" Debate

This is an active area of methodological discussion in the mixed models literature.

### The "Maximal" Philosophy (Barr et al., 2013)

**Argument:** Include random slopes for all within-subject predictors to avoid anti-conservative inference (inflated Type I error).

**Rationale:**

- Failing to include a needed random slope underestimates standard errors
- This leads to false positives for fixed effects
- Better to be conservative (include random slopes) than anti-conservative

**Quote:** "Keep it maximal: Random slopes generalize better and prevent false positives."

### The "Parsimonious" Philosophy (Bates et al., 2015; Matuschek et al., 2017)

**Argument:** Overly complex random effects structures cause:

- Convergence failures
- Singular fits (variance estimates at boundary)
- Unstable estimates
- Loss of statistical power

**Recommendation:**

- Start with a maximal structure
- Simplify by removing variance components that are near zero
- Use model selection criteria (AIC, BIC) or LRT
- Balance complexity with model stability

### Practical Guidelines

| Situation | Recommendation |
|-----------|----------------|
| Large sample size, few random effects | Try maximal structure |
| Small sample size, many predictors | Start parsimonious, add complexity carefully |
| Convergence issues | Simplify random effects structure |
| Singular fit warnings | Remove near-zero variance components |
| Key hypothesis involves slope variation | Include that random slope regardless |

### A Balanced Approach

1. **Start with theory:** What random effects make substantive sense?
2. **Visualize:** Do spaghetti plots suggest slope heterogeneity?
3. **Test:** Use LRT (with boundary corrections) to assess need
4. **Check convergence:** Ensure model fits properly
5. **Report sensitivity:** Show results with/without random slopes if uncertain

***

## Quick Reference Summary

### Deciding on Random Slopes

- **Theory suggests varying slopes?**
  - YES: Include random slope
  - NO: Check diagnostics
    - Fanning in residuals? Add random slope
    - No fanning? RI-only may suffice
    - Uncertain? Fit both, compare via LRT (ML)

### Key R Commands

```{r}
#| eval: false

# Fit models
lme(y ~ time, random = ~time|id, data = mydata)           # nlme
lmer(y ~ time + (1 + time|id), data = mydata)             # lme4

# Extract variance components
VarCorr(model)

# BLUPs
ranef(model)                   # Random effects
coef(model)                    # Subject-specific coefficients

# Predictions
predict(model, level = 0)      # Population average (nlme)
predict(model, level = 1)      # Subject-specific (nlme)
predict(model, re.form = NA)   # Population average (lme4)
predict(model, re.form = NULL) # Subject-specific (lme4)

# Model comparison (use ML!)
anova(fit_ri_ml, fit_rs_ml)
```

### Variance Component Interpretation

| Parameter | Meaning |
|-----------|---------|
| $\sigma^2_{b_0}$ | Between-subject variance at reference time |
| $\sigma^2_{b_1}$ | Between-subject variance in slopes |
| $\rho_{01}$ | Correlation between intercepts and slopes |
| $\sigma^2$ | Within-subject (residual) variance |

***

## References

- Barr, D. J., Levy, R., Scheepers, C., & Tily, H. J. (2013). Random effects structure for confirmatory hypothesis testing: Keep it maximal. *Journal of Memory and Language, 68*(3), 255-278.

- Bates, D., Kliegl, R., Vasishth, S., & Baayen, H. (2015). Parsimonious mixed models. *arXiv preprint arXiv:1506.04967*.

- Matuschek, H., Kliegl, R., Vasishth, S., Baayen, H., & Bates, D. (2017). Balancing Type I error and power in linear mixed models. *Journal of Memory and Language, 94*, 305-315.

- Pinheiro, J. C., & Bates, D. M. (2000). *Mixed-Effects Models in S and S-PLUS*. Springer.

- Singer, J. D., & Willett, J. B. (2003). *Applied Longitudinal Data Analysis*. Oxford University Press.
