---
title: "BIOS 667 - Lecture N: [Topic] (Ch. N)"
subtitle: "Fitzmaurice, Laird & Ware (2011) - Applied Longitudinal Data Analysis"
author: "Naim Rashid"
format:
  revealjs:
    theme: [default]
    footer: "BIOS 667 - UNC Gillings - Lecture N (Ch.N)"
    slide-number: true
    hash: true
    toc: false
    code-overflow: wrap
    code-line-numbers: false
    math: mathjax
    incremental: true
    embed-resources: true
    chalkboard: false
    css: "../../unc-gillings.css"
    scrollable: true
execute:
  warning: false
  message: false
---

<!--
================================================================================
BIOS 667 REFERENCE DECK TEMPLATE (2026 traditional track). See 2026/REFERENCE_DECK_SPEC.md.
Complete, book-faithful, lecture-only; may span 2-3 periods; NO in-class evaluation segments.
DETAIL BAR: "teachable cold" by a substitute (David). The scaffolding below (instructor
notes, timing, prerequisites/recap, notation, case study, pitfalls, datasets/packages,
objectives->assessment, further reading) is what makes that possible.
Per-slide instructor notes go in `::: {.notes}` blocks (presenter view, press S) and are
EXPECTED on concept/example/diagnostic slides. Title style "BIOS 667 - Lecture N: Topic (Ch. N)";
no em-dashes. Verify every formula/claim against FLW (textbook corpus + physical TOC).
Replace every [BRACKETED] placeholder.

WRITING VOICE (do not let slides read as AI-generated):
- NO em-dashes (-) anywhere. Use a period, a comma, or parentheses instead.
- NO semicolons unless genuinely joining two independent clauses; prefer two short sentences.
- Avoid AI-tells: "delve", "leverage" (as a verb), "it's worth noting", "in today's world",
  "a powerful tool", "the world of", hedge stacks ("it is important to note that..."), and
  three-item lists padded for rhythm. Write the way you would say it to the class.
- Short declarative sentences. Cut filler. One idea per bullet. Prefer concrete over abstract.
================================================================================
-->

## Lecture Objectives

By the end of this lecture, you will be able to:

- [measurable objective 1]
- [measurable objective 2]
- [measurable objective 3]

::: {.callout-note}
**Reading.** FLW Ch. N (sections [a-b]). **Assessed by:** [HW#, the final-project skill this sets up].
:::

## Roadmap and pacing

Classes are **75 minutes** (about 25-40 slides each). This lecture spans **[1-3] class periods**
(per `LECTURE_TIMING_PLAN.md`); each Part below is one 75-minute block, split at a natural FLW seam.

- **Part I (Day 1, ~75 min):** [theme]
- **Part II (Day 2, ~75 min):** [theme]
- **Part III (Day 3):** [only if warranted; delete this line otherwise]

<!-- One period only? Delete the Part dividers and this multi-day list. If editing pushes the deck
     past ~40 slides / 75 min, add a `# Part N (Day N): ...` divider and reflow, do not compress. -->

## Prerequisites / recall

Before this lecture, students should be able to:

- [prior skill 1, from Ch. [X] / Lecture [N-1]]
- [prior skill 2]

::: {.notes}
Instructor: confirm the recall items in one minute; if shaky, point to Lecture [N-1] rather than
re-teaching. [What students most often forget here.]
:::

<!-- Notation: include the shared canonical slide. Do NOT paste a per-deck table.
     Source of truth: 2026/NOTATION_REFERENCE.md. Add lecture-specific symbols on a
     small slide AFTER this include, not by editing the shared partial.
     NO-COLLISION RULE: never reuse a reserved canonical symbol for something else.
     X_i / beta = fixed design and coefficients; Z_i / b_i = random-effect design and
     random effects ONLY (e.g. do not write a fixed covariate as z_{ij}). -->
{{< include _notation_box.qmd >}}

# Part I (Day 1): [Theme] {.center background-color="#13294B"}

## Motivation

[Why this topic; the problem it solves; connection to prior lectures ("recall Ch. X ...").]

::: {.notes}
Instructor: [the one-sentence hook]; [why students should care]; transition to the model.
:::

## [Concept / model]

$$
[\text{display the model / key equation, standard notation}]
$$

- [interpretation point]
- [assumption]

::: {.notes}
Instructor: derive [step] on the board; emphasize [the key idea]; the common error is [X]. Tie
back to the notation slide. [Anticipated question + answer.]
:::

## Setup (data and packages)

```{r}
#| label: setup
#| include: false
# Packages used in this lecture: [list, e.g. nlme, lme4, ggplot2, emmeans].
# Load the FLW dataset with a download-fallback path (see CLAUDE.md), so it runs cold.
```

## Case study: [FLW dataset]

[Data description: source, design, variables, research question. This is the formal worked
example on real FLW data, matching the chapter's "Case Studies" section.]

```{r}
#| label: case-fit
#| echo: true
#| output-location: slide
# Model specification on the real FLW dataset.
```

**Interpretation:** [plain-language reading of the estimates, in context, with units].

::: {.notes}
Instructor: walk the output line by line; the number to highlight is [X]; what "good" looks like;
[what students misread here].
:::

## The same model in SAS

R is the course's runnable language (shown above and used in homework). For students who use SAS,
here is the equivalent, shown for reference only (not executed). Mirrors FLW's "Computing" section.

```sas
/* Not run here. SAS PROC equivalent of the R model above. */
PROC MIXED DATA=[dataset] METHOD=REML;
  CLASS id;
  MODEL [y] = [fixed effects] / SOLUTION;
  RANDOM [random effects] / SUBJECT=id TYPE=[UN|VC] G;
RUN;
```

::: {.notes}
Instructor: point out the R-to-SAS mapping (RANDOM/SUBJECT = the `(... | id)` term; METHOD=REML =
the default). Do not run it; it is for SAS users' reference.
:::

::: {.callout-note collapse="true"}
## Check Your Understanding
[1-2 conceptual questions with answers. NON-graded pedagogy.]
:::

# Part II (Day 2): [Theme] {.center background-color="#13294B"}

## Recap of Part I

- [the model and what we concluded last period, in 2-3 bullets]

## [Extension / estimation]

[Continue the chapter; second coherent teaching block.]

## Diagnostics

[Show and ANNOTATE diagnostic plots: what "good" vs "problematic" looks like, with reference
lines and what they mean.]

::: {.notes}
Instructor: [which plot to read first]; the pattern that signals [problem]; the fix is [Y].
:::

## [Testing / model selection]

[LRT/Wald/score contexts; boundary/mixture caveat for variance-component tests; AIC/REML notes.]

## Common pitfalls and misconceptions

- [mistake 1 and why it is wrong]
- [mistake 2]

# Wrap-up {.center background-color="#13294B"}

## Summary / Takeaways

- [key point 1]
- [key point 2]
- [key point 3]

## Recipe card: [workflow]

1. [step]
2. [step]
3. [step]

## Reporting checklist

- [ ] [estimates with 95% CI, variance components, diagnostics, software/method, ...]

## Further reading and for next time

- FLW Ch. N "Further Reading"; [key citations].
- Next: FLW Ch. [N+1]; [forward pointer].
