---
title: "BIOS 667 - Lecture 5: Modeling the Mean - Response Profiles (Ch. 5)"
subtitle: "Fitzmaurice, Laird & Ware (2011) - Applied Longitudinal Data Analysis"
author: "Naim Rashid"
format:
  revealjs:
    theme: [default]
    footer: BIOS 667 · UNC Gillings - Lecture 5 (Ch.5)
    slide-number: true
    hash: true
    toc: false
    code-overflow: wrap
    code-line-numbers: false
    math: mathjax
    incremental: true
    embed-resources: true
    chalkboard: false
    scrollable: true
    css: "../../unc-gillings.css"
execute:
  echo: false
  warning: false
  message: false
---

## Lecture Objectives

By the end of this lecture, you will be able to:

- Define response profiles and explain when they are appropriate for longitudinal data
- Formulate and test the three key hypotheses: parallelism, flatness, and same level
- Set up the general linear model (LM) framework for response profile analysis
- Choose among four baseline handling strategies based on study design
- Implement response profile analysis in R using `nlme::gls()` with unstructured covariance

---

## Roadmap

| Section | Topic |
|---------|-------|
| 1 | Motivation and concept of response profiles |
| 2 | Three hypotheses concerning profiles |
| 3 | General linear model (LM) setup and design matrices |
| 4 | Baseline value strategies |
| 5 | Strengths and weaknesses |
| 6 | R computing example (TLC trial) |

**Pacing:** Spans 2 class periods. Part I (Day 1): Sections 1-3 (concept, hypotheses, LM setup), about 40-45 min. Part II (Day 2): Sections 4-6 (baseline strategies, strengths, computing), about 45-50 min.

---

## Recall from Lecture 4 (Ch. 4)

Response profiles are the first place we **use** the estimation and inference machinery from Lecture 4:

- **GLS / ML** estimation of $\boldsymbol\beta$ given a covariance $\Sigma_i$ (here left unstructured).
- **REML** for the covariance parameters; **ML** when we compare nested mean models by likelihood-ratio test.
- The **LRT / Wald** tests, which we use below to test parallelism, flatness, and group differences.

So this lecture is "Chapter 4 applied to the mean response over time."

::: {.notes}
- One-slide bridge. Remind them Lecture 4 gave the tools (GLS/ML, REML vs ML, the LRT); this lecture is the first real application to a scientific question.
- The key carry-over: fit the covariance by REML, but refit competing mean models by ML before a likelihood-ratio test.
:::

---

{{< include _notation_box_intro.qmd >}}

## Notation for this lecture (Ch. 5)

::: {.callout-note .notation-card}
## Counts and indices (FLW convention)

Response profiles assume a **balanced** design: $N$ subjects, each measured at the **same** $n$ occasions, in $G$ groups.

| Symbol | Meaning |
|:--|:--|
| $N$ | number of **subjects** ($i = 1,\dots,N$) |
| $n$ | number of **occasions**, common to all subjects ($j = 1,\dots,n$) |
| $G$ | number of **groups** ($g = 1,\dots,G$ indexes group) |
| $\Sigma_i$ | marginal covariance of $\mathbf{Y}_i$: $\Sigma_i = \operatorname{Cov}(\mathbf{Y}_i)$ (FLW's letter) |

This is the standard FLW convention ($N$ subjects, $n$ occasions); the running notation card uses the same $N$ for subjects and $n_i$ for occasions, which here is the constant $n$. Ch. 5 is purely marginal (no random effects), so $\Sigma_i$ is the full (unstructured) covariance of $\mathbf{Y}_i$, exactly FLW's $\operatorname{Cov}(\mathbf{Y}_i\mid\mathbf{X}_i)=\Sigma_i$; the deck sometimes drops the subscript and writes $\Sigma$. Reserved symbols ($X$, $\boldsymbol\beta$, $Z$, $b$) keep their usual roles.
:::

::: {.callout-note}
## Sub/superscript key
Superscript $(g)$ = group $g$; subscript $j$ = occasion $j$. So $\mu_j^{(g)}$ is the mean at occasion $j$ in group $g$.
:::

# Part I (Day 1): Concept, Hypotheses, and the LM Setup

::: {.notes}
- Day 1 covers Sections 1-3: the concept of response profiles, the three hypotheses, and the general linear model (LM) design.
- Pace: about 40-45 minutes. Sections 1-2 are quick (concept + hypotheses); spend the back half on the LM design and contrasts.
- Goal by end of Day 1: students can state parallelism / flatness / same-level formally and write the cell-means design matrix.
:::

# Section 1: Motivation and Concept

## Why Response Profiles?

**Goal:** Analyze longitudinal data with minimal structure on:

- Mean response trajectory
- Covariance among repeated measures

---

## What Are Response Profiles?

Response profiles are sequences of group-specific mean responses over time.

**Key features:**

- Applies to **balanced longitudinal designs**
- Same measurement occasions for all subjects
- Can accommodate **missing data** (balanced but incomplete designs)

---

## Computing Group-Specific Means

Suppose groups are defined by treatment or observational factors.

At each time $t_j$, compute the group-specific mean:

$$
\hat{\mu}_g(t_j) = \frac{1}{n_g} \sum_{i \in g} Y_{ij}
$$

Plotting $\hat{\mu}_g(t_j)$ over time yields the **mean response profile**.

::: {.callout-note}
Profiles capture *patterns of change* rather than imposing parametric structure.
:::

---

## Example: TLC Trial Data

```{r}
#| echo: true
#| output-location: slide
# Published TLC group means from FLW Table 5.6; full long data loaded later (Section 6)
time <- c(0, 1, 4, 6)
succimer <- c(26.5, 13.5, 15.5, 20.8)
placebo  <- c(26.3, 24.7, 24.1, 23.6)

plot(time, succimer, type="b", col="blue", pch=19, ylim=c(10,30),
     xlab="Time (weeks)", ylab="Mean Blood Lead (µg/dL)",
     main="TLC group means (FLW Table 5.6)")
lines(time, placebo, type="b", col="red", pch=17)
legend("topright", legend=c("Succimer", "Placebo"),
       col=c("blue","red"), pch=c(19,17))
```

::: {.notes}
- This is the motivating picture for the whole lecture: two group mean profiles over four occasions.
- Emphasize: succimer drops sharply at week 1, then rebounds; placebo stays flat. The profiles are clearly NOT parallel.
- These are the actual published TLC group means from FLW Table 5.6, not synthetic numbers; we load and fit the full TLC trial in Section 6.
- Common student question: "why not just fit a line?" Answer: we are imposing no functional form, just cell means per occasion.
- Transition: the eye sees non-parallel profiles. Next we make "parallel" a formal, testable hypothesis.
:::

---

## TLC Trial Observations

| Group | Pattern |
|-------|---------|
| Succimer | Sharp decline post-baseline, then rebound |
| Placebo | Remains relatively stable over time |

---

## Discussion: Handling Baseline

Baseline measurements are special:

- Typically pre-randomization in trials (independent of treatment)
- May be range-restricted or differ systematically in observational studies

---

## Key Question: How Should Baseline Be Handled?

| Option | Description |
|--------|-------------|
| Include in profile | Treat baseline as part of the response vector |
| Adjust explicitly | Use ANCOVA or contrasts |

Implications for hypothesis testing, efficiency, and power.

---

## Think-Pair-Share

::: {.callout-tip}
**Discussion questions:**

1. In a randomized trial, why is it reasonable to treat baseline as independent of treatment?
2. How might this differ in an observational study?
:::

---

::: {.callout-note collapse="true"}
## Check Your Understanding: Response Profiles Basics

1. What type of longitudinal design is response profile analysis best suited for?
2. Why do we call them "response profiles" rather than "response curves"?
3. In the TLC trial plot, what does the sharp drop in the Succimer group at week 1 suggest?

**Answers:**

1. **Balanced longitudinal designs** where all subjects have the same planned measurement occasions (though missing data is allowed)
2. Because we estimate **discrete means at each time point** rather than assuming a continuous functional form (e.g., linear, quadratic)
3. The treatment has an immediate, strong effect on reducing blood lead levels, but the rebound suggests the effect may not be sustained
:::

# Section 2: Hypotheses Concerning Profiles

## Three Main Hypotheses

Given $n$ occasions (time points) across $G$ groups, we test:

| Hypothesis | Description | Degrees of Freedom |
|------------|-------------|-------------------|
| $H_{01}$ | Parallelism (no group $\times$ time interaction) | $(G-1)(n-1)$ |
| $H_{02}$ | Flatness (no time effect) | $n-1$ |
| $H_{03}$ | Same level (no group effect) | $G-1$ |

::: {.callout-note}
$G$ = number of groups, $n$ = number of occasions (Ch. 5 convention). Tiny example: $G=2$ groups, $n=4$ occasions gives parallelism df $=(G-1)(n-1) = 1 \times 3 = 3$.
:::

::: {.notes}
- These three hypotheses are the entire inferential agenda for response profiles. Write them on the board before showing the table.
- Re-state the Ch. 5 convention out loud: here $n$ is OCCASIONS, $N$ is subjects. This is the single most common source of confusion in this chapter.
- Walk the df: parallelism counts the free group-by-time interaction contrasts, $(G-1)(n-1)$; flatness is $n-1$ time contrasts; same-level is $G-1$ group contrasts.
- Common student error: reading $n$ as the sample size. Plug in $G=2, n=4$ on the board to anchor it.
- Transition: parallelism is the gatekeeper test, so we cover the testing order next.
:::

---

## Testing Order Principle

**If $H_{01}$ is rejected**, main effects ($H_{02}$, $H_{03}$) are not directly interpretable due to the presence of interaction.

**Takeaway:** Test parallelism first; only then consider main effects.

---

## Randomized vs. Observational Studies

| Study Type | Baseline Property | Primary Focus |
|------------|-------------------|---------------|
| Randomized trial | Groups equal at baseline by design | $H_{01}$ (patterns of change) |
| Observational | Groups may differ at baseline | $H_{02}$, $H_{03}$ meaningful if $H_{01}$ holds |

---

## Visual: Parallel Profiles (No Interaction)

Two groups with **parallel** mean trajectories (constant difference over time).

```{r}
#| echo: true
#| output-location: slide
set.seed(1)
t <- 1:5
mu1 <- 10 + 2*t                # Group 1
mu2 <- 12 + 2*t                # Group 2 (constant +2 offset)

plot(t, mu1, type="b", pch=19, ylim=range(c(mu1, mu2)),
     xlab="Time", ylab="Mean Response", main="(a) Parallel Mean Profiles")
lines(t, mu2, type="b", pch=17)
abline(v=t, col="gray90", lty=3)
legend("topleft", legend=c("Group 1","Group 2"), pch=c(19,17), bty="n")
```

---

## Visual: Flat Profiles (No Time Effect)

Assuming parallelism holds, flat profiles test:

$$
H_{02}: \mu_1 = \mu_2 = \cdots = \mu_n
$$

```{r}
#| echo: true
#| output-location: slide
t <- 1:5
mu1 <- rep(15, length(t))
mu2 <- rep(20, length(t))

plot(t, mu1, type="b", pch=19, ylim=range(c(mu1, mu2)),
     xlab="Time", ylab="Mean Response", main="(b) Flat Mean Profiles (No Time Effect)")
lines(t, mu2, type="b", pch=17)
abline(v=t, col="gray90", lty=3)
legend("topleft", legend=c("Group 1","Group 2"), pch=c(19,17), bty="n")
```

---

## Visual: Same Level (No Group Effect)

Assuming parallelism holds, same level tests:

$$
H_{03}: \boldsymbol{\mu}^{(1)} = \boldsymbol{\mu}^{(2)} = \cdots = \boldsymbol{\mu}^{(G)}
$$

```{r}
#| echo: true
#| output-location: slide
t <- 1:5
mu  <- 10 + 1.5*t

# The two group profiles coincide by construction (no group effect);
# add a tiny vertical jitter so both series are visible.
plot(t, mu, type="b", pch=19, ylim=range(mu),
     xlab="Time", ylab="Mean Response", main="(c) Same Level Across Groups (No Group Effect)")
lines(t, mu + 0.1, type="b", pch=17, col="gray40")
abline(v=t, col="gray90", lty=3)
legend("topleft", legend=c("Group 1","Group 2 (jittered)"), pch=c(19,17),
       col=c("black","gray40"), bty="n")
```

---

## Formal Statement: Two Groups

For Treatment ($T$) vs. Control ($C$), define:

$$
\Delta_j = \mu_j^{(T)} - \mu_j^{(C)}
$$

Test parallelism:

$$
H_{01}: \Delta_1 = \Delta_2 = \cdots = \Delta_n \quad (n-1 \text{ df})
$$

::: {.notes}
- This is the formal definition of parallelism: define the group gap $\Delta_j$ at each occasion, then ask whether that gap is constant across occasions.
- Emphasize the picture-to-algebra bridge: "parallel profiles" means "$\Delta_j$ does not depend on $j$".
- With two groups parallelism has $n-1$ df: there are $n$ gaps but the null sets them all equal, leaving $n-1$ free comparisons.
- Common student error: thinking $\Delta_j = 0$ is the null. It is not; the gap can be large, it just has to be the SAME at every occasion.
- Transition: with more than two groups we carry a $\Delta$ per non-reference group, which multiplies the df.
:::

---

## Formal Statement: More Than Two Groups

With reference group $G$, define:

$$
\Delta_j^{(g)} = \mu_j^{(g)} - \mu_j^{(G)}
$$

Test parallelism:

$$
H_{01}: \Delta_1^{(g)} = \cdots = \Delta_n^{(g)} \quad \forall g=1,\dots,G-1
$$

with $(G-1)(n-1)$ degrees of freedom.

---

::: {.callout-note collapse="true"}
## Check Your Understanding: Profile Hypotheses

1. If you have 3 groups and 5 time points, how many degrees of freedom does the parallelism test have?
2. Why should you test parallelism ($H_{01}$) before interpreting main effects?
3. If profiles are parallel, what does rejecting $H_{02}$ (flatness) tell you?

**Answers:**

1. $(G-1)(n-1) = (3-1)(5-1) = 8$ degrees of freedom
2. If there is a group $\times$ time interaction (non-parallel profiles), the "group effect" and "time effect" vary depending on which time/group you look at - main effects are not meaningful without specifying the level of the other factor
3. Rejecting flatness means there **is** a time effect - the mean response changes over time (at least one time point differs from the others)
:::

# Section 3: General Linear Model (LM) Setup

## The General Linear Model (LM) for Response Profiles

Represent mean response profiles using the general (normal-errors) linear model, the **LM**:

$$
E(\mathbf{Y}_i \mid \mathbf{X}_i) = \boldsymbol{\mu}_i = \mathbf{X}_i \boldsymbol{\beta}
$$

**Notation (Ch. 5):**

| Symbol | Meaning |
|--------|---------|
| $n$ | Number of occasions (time points) |
| $N$ | Number of subjects |
| $G$ | Number of groups |
| Total parameters | $G \times n$ |

::: {.callout-note}
"LM" here is the **general** linear model with normal errors. It is not the **generalized** linear model (GLM), which we reach in Ch. 11.
:::

::: {.notes}
- Anchor the whole section: the cell means $\mu_j^{(g)}$ are just regression coefficients in an LM, so all our machinery (contrasts, Wald, LRT) applies.
- Be explicit that "LM" = general linear model with normal errors; reserve "GLM" for the generalized linear model in Ch. 11. Students conflate these constantly.
- Re-state the count: $G$ groups times $n$ occasions gives $G \times n$ free cell means in the saturated mean model.
- Common student error: treating $X_i$ as a fixed design across subjects. It is subject-specific; rows drop out when occasions are missing.
- Transition: make it concrete with two groups and three occasions.
:::

---

## Example: Two Groups, Three Occasions

Mean response profiles (column vectors):

$$
\boldsymbol{\mu}^{(1)} =
\begin{pmatrix} \mu_1^{(1)} \\ \mu_2^{(1)} \\ \mu_3^{(1)} \end{pmatrix}, \quad
\boldsymbol{\mu}^{(2)} =
\begin{pmatrix} \mu_1^{(2)} \\ \mu_2^{(2)} \\ \mu_3^{(2)} \end{pmatrix}
$$

Parameter vector:

$$
\boldsymbol{\beta} = (\beta_1, \beta_2, \ldots, \beta_6)^\top
$$

---

## Design Matrix Example

For a subject in **group 1** with all three occasions (cell-means coding):

$$
\mathbf{X}_i =
\begin{bmatrix}
1 & 0 & 0 & 0 & 0 & 0 \\
0 & 1 & 0 & 0 & 0 & 0 \\
0 & 0 & 1 & 0 & 0 & 0
\end{bmatrix}
$$

Columns 1-3 are the **group-1 cell means** (occasions 1, 2, 3); columns 4-6 are the **group-2 cell means**. So $\mathbf{X}_i \boldsymbol{\beta}$ picks out group 1 here. The map is:

$$
\beta_1 = \mu_1^{(1)},\ \beta_2 = \mu_2^{(1)},\ \beta_3 = \mu_3^{(1)},\quad
\beta_4 = \mu_1^{(2)},\ \beta_5 = \mu_2^{(2)},\ \beta_6 = \mu_3^{(2)}
$$

::: {.notes}
- This is the workhorse design: each column is one group-by-occasion cell mean, so $\boldsymbol{\beta}$ IS the vector of cell means. No intercept-plus-contrasts to decode.
- Walk it row by row: a group-1 subject's row $j$ has a single 1 in the group-1 occasion-$j$ column, so $E(Y_{ij}) = \beta_j$.
- Point at columns: 1-3 are group 1's three means, 4-6 are group 2's three means. A group-2 subject would have 1s in columns 4-6 instead.
- Common student error: expecting an intercept column. Cell-means coding has none; every column is a mean.
- Transition: hypotheses about parallelism are now just linear restrictions on this $\boldsymbol{\beta}$.
:::

---

## Expressing Hypotheses via Contrasts

Hypothesis of no group $\times$ time interaction:

$$
H_{01}: \mathbf{L}\boldsymbol{\beta} = \mathbf{0}
$$

Each row of $\mathbf{L}$ encodes one $\Delta_j - \Delta_{j+1} = 0$ comparison, where $\Delta_j = \mu_j^{(2)} - \mu_j^{(1)}$. With $\boldsymbol{\beta} = (\mu_1^{(1)},\mu_2^{(1)},\mu_3^{(1)},\mu_1^{(2)},\mu_2^{(2)},\mu_3^{(2)})^\top$, one row testing $\Delta_1 = \Delta_2$ is:

$$
\mathbf{L}_{\text{row 1}} = \begin{pmatrix} -1 & 1 & 0 & 1 & -1 & 0 \end{pmatrix}
\;\Rightarrow\;
\mathbf{L}_{\text{row 1}}\boldsymbol{\beta} = (\mu_2^{(2)} - \mu_2^{(1)}) - (\mu_1^{(2)} - \mu_1^{(1)}) = \Delta_2 - \Delta_1
$$

Setting all $n-1$ such rows to $\mathbf{0}$ forces $\Delta_1 = \cdots = \Delta_n$ (parallelism).

**Testing methods:** Wald or likelihood ratio tests

::: {.notes}
- The payoff slide: every hypothesis from Section 2 is a row-set of $\mathbf{L}$ on the cell-means $\boldsymbol{\beta}$.
- Walk the single row out loud: $\mathbf{L}_{\text{row 1}}\boldsymbol{\beta}$ equals $\Delta_2 - \Delta_1$, so setting it to zero says the gap at occasion 2 equals the gap at occasion 1.
- Stack $n-1$ such rows for the full parallelism test, matching the $(G-1)(n-1)$ df from Section 2 (here $G=2$).
- Common student error: sign bookkeeping. Show that the group-1 entries and group-2 entries have opposite signs because $\Delta_j$ is a difference.
- Transition: Wald uses the fitted $\hat{\boldsymbol{\beta}}$ and its covariance; LRT refits the reduced model. We use the LRT in R later.
:::

---

## Visual: Observed Means

```{r}
#| echo: true
#| output-location: slide
time <- 1:3
mu_group1 <- c(10, 14, 18)
mu_group2 <- c(11, 16, 21)  # slightly different slope

plot(time, mu_group1, type="b", pch=19, col="blue", ylim=c(10,22),
     xlab="Time", ylab="Mean Response", main="Observed Means")
lines(time, mu_group2, type="b", pch=17, col="red")
legend("topleft", legend=c("Group 1","Group 2"),
       col=c("blue","red"), pch=c(19,17), bty="n")
```

---

## Visual: Fitted Full Model

Full model: each group $\times$ time cell estimated freely.

```{r}
#| echo: true
#| output-location: slide
plot(time, mu_group1, type="b", pch=19, col="blue", ylim=c(10,22),
     xlab="Time", ylab="Mean Response", main="Fitted Means (Full Model)")
lines(time, mu_group2, type="b", pch=17, col="red")
legend("topleft", legend=c("Group 1","Group 2"),
       col=c("blue","red"), pch=c(19,17), bty="n")
```

---

## Visual: Fitted Under the Null

Null hypothesis: group difference $\Delta$ is constant over time.

```{r}
#| echo: true
#| output-location: slide
delta <- mean(mu_group2 - mu_group1)   # average difference
mu_group2_null <- mu_group1 + delta

plot(time, mu_group1, type="b", pch=19, col="blue", ylim=c(10,22),
     xlab="Time", ylab="Mean Response", main="Fitted Means (Null: Parallel Profiles)")
lines(time, mu_group2_null, type="b", pch=17, col="red", lty=2)
legend("topleft", legend=c("Group 1","Group 2 (Null Fit)"),
       col=c("blue","red"), pch=c(19,17), lty=c(1,2), bty="n")
```

---

## Handling Missing Data

**Advantage of the LM formulation:** Flexible design matrix $\mathbf{X}_i$.

If a subject is missing a time point, remove the corresponding row from $\mathbf{X}_i$.

Still valid to fit model using available observations.

---

## Reference Group Parameterization

Alternative coding: choose one group as **reference**.

With $G$ groups, define group-indicator covariates $W_{ig}$ (the symbol $Z$ is reserved for random effects, Ch. 8):

$$
W_{ig} =
\begin{cases}
  1 & \text{if subject } i \text{ belongs to group } g \\
  0 & \text{otherwise}
\end{cases}
$$

---

## Reference Group Interpretation

If we set $\mathbf{X}_i = (1, W_{i1}, \ldots, W_{i,G-1})$, then:

$$
\mu^{(g)} =
\begin{cases}
  \beta_1 + \beta_{g+1}, & g=1,\ldots,G-1 \\
  \beta_1, & g=G \quad (\text{reference group})
\end{cases}
$$

| Parameter | Interpretation |
|-----------|----------------|
| $\beta_1$ | Mean of reference group |
| $\beta_{g+1}$ | Deviation of group $g$ from reference |

---

## R Example: Reference Coding

```{r}
#| echo: true
#| output-location: slide
# Example: 3 groups, using Group 3 as reference
group <- factor(c("G1","G2","G3","G1","G2","G3"))

# Default coding in R: last group is reference
X <- model.matrix(~ group)
X
```

---

## Reference Coding Interpretation

- Intercept corresponds to Group 3 mean
- Columns `groupG1` and `groupG2` are deviations from Group 3
- This is the default parameterization in R

---

## Testing Main Effects

If profiles are parallel ($H_{01}$ not rejected):

| Test | Hypothesis | Degrees of Freedom |
|------|------------|-------------------|
| No time effect | $H_{02}: \mu_1 = \mu_2 = \cdots = \mu_n$ | $n-1$ |
| No group effect | $H_{03}: \boldsymbol{\mu}^{(1)} = \cdots = \boldsymbol{\mu}^{(G)}$ | $G-1$ |

Both tested using a reduced model (main effects only, no interaction).

::: {.callout-note}
The flat-profile and same-level pictures are the Section 2 visuals; we do not repeat them here.
:::

---

## LM Framework: Key Takeaways

- Encodes group $\times$ time means in $\boldsymbol{\beta}$
- Tests hypotheses via linear contrasts ($\mathbf{L}\boldsymbol{\beta} = \mathbf{0}$)
- Full model: exact means per cell
- Null model: imposes parallelism (constant group difference)

# Part II (Day 2): Baseline Strategies, Strengths, and Computing

::: {.notes}
- Day 2 covers Sections 4-6: baseline-handling strategies, strengths and weaknesses, and the R/SAS computing on the TLC trial.
- Pace: about 45-50 minutes. The baseline strategies and the change-vs-ANCOVA efficiency argument deserve the most time; the computing section is demonstration.
- Recap Day 1 in one line: response profiles are cell means in an LM, tested by contrasts on $\boldsymbol{\beta}$, parallelism first.
- Goal by end of Day 2: students can pick a baseline strategy by design and run the gls() fit and the ANCOVA in R.
:::

# Section 4: Baseline Value Strategies

## Four Strategies for Baseline Handling

| Strategy | Description |
|----------|-------------|
| 1 | Retain baseline; no equality assumption |
| 2 | Retain baseline; constrain groups equal at baseline (RCT) |
| 3 | Analyze change-from-baseline: $\Delta_{ij} = Y_{ij} - Y_{i1}$ |
| 4 | ANCOVA: use $Y_{i1}$ as covariate for post-baseline responses |

::: {.notes}
- This is the map for all of Section 4: four ways to treat the baseline measurement, from "leave it in the response vector" to "use it as a covariate".
- Frame the spine: strategies 1-2 keep baseline as a response; strategies 3-4 use it to define or adjust the outcome.
- The design drives the choice: strategy 2's equal-baseline constraint is only honest under randomization; strategy 4 (ANCOVA) is the efficient default in RCTs.
- Common student error: treating change scores (3) and ANCOVA (4) as the same thing. They answer different questions (unconditional change vs. baseline-conditional difference) and differ in efficiency.
- Transition: take them one at a time, with pros and cons, then quantify the efficiency gap.
:::

---

## Strategy 1: Retain Baseline, No Equality

Fit the LM with all times, no constraint at $t=1$:

$$
E(\mathbf{Y}_i \mid \mathbf{X}_i) = \mathbf{X}_i \boldsymbol{\beta}
$$

where $\mu_1^{(1)}, \ldots, \mu_1^{(G)}$ are free parameters.

---

## Strategy 1: Pros and Cons

| Pros | Cons |
|------|------|
| Minimal assumptions | Main "group" effect hard to interpret if baseline levels differ |
| Natural for observational data | May be less efficient than approaches exploiting baseline balance |
| Uses all information | |

---

## Strategy 2: Constrain Equal at Baseline

Appropriate in **randomized trials** (baseline pre-randomization).

Add constraint:

$$
\mu_1^{(1)} = \mu_1^{(2)} = \cdots = \mu_1^{(G)}
$$

---

## Strategy 2: Pros and Cons

| Pros | Cons |
|------|------|
| Reflects randomization | Invalid if baseline equality fails |
| Often yields greater precision | Sensitivity to constraint violation |
| Clean interpretation of group differences | |

---

## Strategy 3: Change-from-Baseline

Define subject-level differences:

$$
\Delta_{ij} = Y_{ij} - Y_{i1}, \quad j = 2, \ldots, n
$$

Or use a univariate summary:

$$
Y_i^* = \frac{1}{n-1}\sum_{j=2}^n (Y_{ij} - Y_{i1})
$$

---

## Strategy 3: Pros and Cons

| Pros | Cons |
|------|------|
| Intuitive change estimand | Can be less efficient than ANCOVA when $\rho < 1$ |
| Avoids modeling baseline as covariate | Forces the baseline coefficient to exactly 1, instead of the optimal value |
| Simpler tests ($G-1$ df) | Loses information about absolute levels |

---

## Strategy 4: ANCOVA

Univariate ANCOVA on average post-baseline response:

$$
Y_i^* = \alpha + \beta \cdot \text{trt}_i + \gamma \cdot Y_{i1} + \varepsilon_i
$$

Or time-specific repeated-measures ANCOVA:

$$
Y_{ij} = \alpha_j + \beta_j \cdot \text{trt}_i + \gamma_j \cdot Y_{i1} + \varepsilon_{ij}, \quad j \ge 2
$$

---

## Strategy 4: Pros and Cons

| Pros | Cons |
|------|------|
| In RCTs, typically more efficient | In observational data, may induce Lord's paradox (next slide) |
| Adjusts for random baseline variability | Requires linearity and homogeneous slopes |
| Smaller SEs for group effect | Measurement error in $Y_{i1}$ can bias $\gamma$ |

---

## Lord's Paradox (beyond FLW Ch. 5)

::: {.callout-warning}
## Instructor extension, not in FLW Ch. 5

**Lord's paradox:** when two groups differ at baseline, change-score analysis and ANCOVA can give **opposite-signed** group effects on the same data. Neither is "wrong"; they answer different questions (unconditional change vs. baseline-conditional difference).
:::

```
            change score says:            ANCOVA says:
   group A  ---- no change (flat) ----     conditional on baseline,
   group B  ---- no change (flat) ----     A and B differ in outcome
   (groups start at different baselines)   (regression line tilts the comparison)
   => "no group effect"                    => "group effect" (opposite conclusion)
```

The two methods point in different directions precisely because the groups are not comparable at baseline (no randomization).

::: {.aside}
Lord, F.M. (1967). A paradox in the interpretation of group comparisons. *Psychological Bulletin* 68(5):304-305.
:::

---

## Relative Efficiency: Change vs. ANCOVA

Under compound symmetry with variance $\sigma^2$ and correlation $\rho$, with **one baseline and one post-baseline occasion** ($n = 2$):

$$
\text{RE} = \frac{\text{Var}(\hat\beta_{\text{Change}})}{\text{Var}(\hat\beta_{\text{ANCOVA}})} = \frac{2}{1 + \rho}
$$

This is the single-follow-up special case. The general $n$-occasion result (FLW eq. (5.3), Section 5.6) gives the efficiency of the change-score summary **relative to ANCOVA** as

$$
\frac{1}{n}\{1 + (n-1)\rho\} < 1,
$$

so ANCOVA is $\dfrac{n}{1 + (n-1)\rho}$ times as efficient (at $n = 2$ this is $2/(1+\rho)$). ANCOVA's advantage grows BOTH as $\rho$ decreases AND as the number of occasions $n$ increases (an $n$-fold gain as $\rho \to 0$).

**Why ANCOVA wins:** a change score writes $Y_f - 1 \cdot Y_1$, forcing the baseline coefficient to exactly **1**. ANCOVA fits $Y_f - \gamma \cdot Y_1$ and **estimates the optimal** $\gamma$ (here $\gamma \approx \rho$), the regression-to-the-mean adjustment. Pinning $\gamma = 1$ is generally not optimal, so change scores carry extra variance.

::: {.notes}
- This is the core conceptual fix: change scores are ANCOVA with the baseline coefficient hard-wired to 1.
- Say it plainly: ANCOVA lets the data choose how much baseline to subtract off (the optimal slope is regression to the mean, roughly $\rho$); change scores subtract all of it.
- Tie to the formula: when $\rho \to 1$ the optimal $\gamma \to 1$ too, so the constraint is harmless and RE $\to 1$; when $\rho$ is low, forcing $\gamma=1$ is far from optimal, so RE $\to 2$.
- Common student error (and an old slide error): blaming "measurement error" or "using baseline twice". That is wrong; the inefficiency is the suboptimal fixed coefficient.
- Transition: read the RE off a small table next.
:::

---

## Relative Efficiency: Implications

The table below is the **single-follow-up ($n = 2$) special case** $\text{RE} = 2/(1+\rho)$:

| Condition | Relative Efficiency ($n = 2$) |
|-----------|---------------------|
| $\rho = 0$ | RE = 2 (change scores twice as variable) |
| $\rho = 0.5$ | RE = 4/3 $\approx$ 1.33 |
| $\rho = 0.8$ | RE = 10/9 $\approx$ 1.11 |
| $\rho \to 1$ | RE $\to$ 1 (methods become equivalent) |

RE $\geq 1$ always, so ANCOVA is always at least as efficient as change scores. With more occasions the gap is larger still: by FLW eq. (5.3) ANCOVA is $n/\{1+(n-1)\rho\}$ times as efficient, up to $n$-fold as $\rho \to 0$.

---

## Decision Guide: When to Use Which?

| Study Type | Recommended Strategy |
|------------|---------------------|
| Randomized trial | Strategy 4 (ANCOVA) or Strategy 2 for power and clarity |
| Observational | Strategy 1 or Strategy 3 for unconditional change questions |
| | Strategy 4 only if conditional estimand is intended |

---

## Estimand Matters

::: {.callout-tip}
- **Strategy 3** tests whether groups differ in mean change (unconditional)
- **Strategy 4** tests whether individuals matched on baseline differ in expected outcome (conditional)
:::

---

## When to Use Alternatives

| Situation | Alternative Approach |
|-----------|---------------------|
| Irregular timing | LMM with random effects; AR(1), SP[POW] |
| Trend focus | Parametric/spline-by-group (Ch. 6) |
| Population-average | GEE (robust SE) |
| Large $n$ vs. $N$ | Structured $\Sigma$ / shrinkage |

---

## Rule of Thumb Summary

| Study Design | Recommendation |
|--------------|----------------|
| RCT | Strategy 2 or 4; Strategy 4 typically more efficient |
| Observational | Strategy 1 for flexible profiling; Strategy 3 for unconditional change; Strategy 4 only when conditional estimand is intended |

---

## Mini Demo: Change vs. ANCOVA

```{r}
#| echo: true
#| output-location: slide
set.seed(7)
n <- 120
group <- factor(rep(c("Control","Treatment"), each=n/2))
Y1 <- rnorm(n, 20, 4)                 # baseline
# True model: follow-up depends on baseline + treatment effect
Yf <- 5 + 0.6*Y1 + 2*(group=="Treatment") + rnorm(n, 0, 3)

# Change score and ANCOVA
change <- Yf - Y1
fit_change <- lm(change ~ group)
fit_ancova <- lm(Yf ~ group + Y1)

summary(fit_change)$coef
summary(fit_ancova)$coef
```

::: {.notes}
- This demo makes the efficiency argument concrete: same data, two estimators of the treatment effect.
- The data are generated with the follow-up regressing on baseline at coefficient 0.6, NOT 1. So the change score's forced coefficient of 1 is wrong, and ANCOVA's estimated coefficient will sit near 0.6.
- Point students at the two treatment-effect rows: both estimate roughly the same effect (about 2), but the ANCOVA SE is smaller.
- Common student error: expecting the point estimates to differ a lot. In a balanced RCT they agree; it is the SE (precision) that separates them.
- Transition: quantify that SE gap and tie it back to RE = 2/(1+rho) on the next slide.
:::

---

## Takeaway: Comparing the Two Treatment-Effect SEs

```{r}
#| echo: true
#| output-location: slide
se_change <- summary(fit_change)$coef["groupTreatment", "Std. Error"]
se_ancova <- summary(fit_ancova)$coef["groupTreatment", "Std. Error"]
rho_hat   <- cor(Y1, Yf)

c(SE_change = se_change, SE_ancova = se_ancova,
  ratio_SE2 = (se_change^2) / (se_ancova^2),   # empirical variance ratio
  RE_formula = 2 / (1 + rho_hat))              # 2/(1+rho)
```

The change-score SE is the larger of the two. This demo has a single follow-up, so the empirical variance ratio tracks the $n = 2$ formula $\text{RE} = 2/(1+\rho)$ (the special case of FLW eq. (5.3)): ANCOVA wins because it estimates the **optimal** baseline coefficient instead of forcing it to 1. The gap is largest at low $\rho$ and vanishes as $\rho \to 1$ (decreasing in $\rho$); with more occasions the gap grows toward $n$-fold.

::: {.notes}
- Close the loop opened on the RE slide: the variance ratio computed from this one dataset should land near 2/(1+rho).
- Same framing as the mechanism slide: ANCOVA estimates the optimal coefficient (regression to the mean); change scores pin it at 1.
- Direction reminder: RE is decreasing in rho, so ANCOVA's advantage is biggest when baseline and follow-up are weakly correlated and disappears as they become identical.
- Common student error: thinking more correlation helps change scores. It helps both, but it shrinks the GAP toward 1.
:::

---

## Baseline Value Strategy Summary

| Strategy | Key Feature |
|----------|-------------|
| 1 | Flexible; good for observational; may be less efficient |
| 2 | Uses RCT baseline equality; more precise if valid |
| 3 | Intuitive changes; simpler tests; may be inefficient |
| 4 | Most powerful in RCTs; in observational data watch for Lord's paradox (defined earlier) |

**Bottom line:** Let design + scientific question dictate the strategy.

---

::: {.callout-note collapse="true"}
## Check Your Understanding: Baseline Strategies

1. In a randomized trial, which baseline strategy is typically most efficient and why?
2. What is the key assumption behind Strategy 2 (constraining groups equal at baseline)?
3. Why might change-from-baseline (Strategy 3) be less efficient than ANCOVA (Strategy 4)?

**Answers:**

1. **Strategy 4 (ANCOVA)** - it adjusts for random baseline variability, reducing residual variance and yielding smaller standard errors for the treatment effect
2. Groups have equal mean response at baseline (valid by randomization in RCTs, but potentially violated in observational studies)
3. A change score is $Y_f - 1 \cdot Y_1$: it forces the baseline coefficient to exactly 1. ANCOVA fits $Y_f - \gamma Y_1$ and estimates the optimal $\gamma$ (the regression-to-the-mean coefficient, about $\rho$). Pinning $\gamma=1$ is generally suboptimal, so change scores carry extra variance. With a single follow-up ($n = 2$) the loss follows $\text{RE} = 2/(1+\rho)$, shrinking to nothing as $\rho \to 1$; in general (FLW eq. (5.3)) ANCOVA is $n/\{1+(n-1)\rho\}$ times as efficient, so its advantage grows as $\rho$ falls and as the number of occasions $n$ grows.
:::

# Section 5: Strengths and Weaknesses

## Strengths of Response Profile Analysis

- Balanced designs with common measurement times
- Arbitrary mean structure + unstructured $\Sigma$
- Avoids covariance misspecification by leaving $\Sigma$ unstructured (no working-correlation assumption)
- Works with balanced-incomplete data
- Discrete covariates via the LM; contrasts/AUC

::: {.callout-note}
Leaving $\Sigma$ free protects against COVARIANCE misspecification; the SEs here are model-based (from the fitted $\Sigma$), not sandwich/empirical. The MEAN model can still be misspecified, so this is not a blanket robustness claim.
:::

---

## Comparison with MANOVA

| MANOVA | Response Profiles |
|--------|-------------------|
| Complete cases only | Handles missing data |
| Test-heavy approach | Estimands + estimation first |
| | Flexible baseline handling |

---

## Weaknesses of Response Profile Analysis

- Requires balanced timing; poor for mistimed visits
- Ignores time ordering; omnibus tests need follow-up
- Lower power for specific trends (prefer 1 df tests)
- Parameter growth: $Gn + \frac{n(n+1)}{2}$

---

## Parameter Growth Example ($G=2$)

```{r}
#| echo: true
#| output-location: slide
param_counts <- function(G, n) G*n + n*(n+1)/2
data.frame(n=c(3,10), total_params=param_counts(2, c(3,10)))
```

# Section 6: R Computing Example

## Computing Goals

- Replicate PROC MIXED response-profiles analysis in R
- Model: $Y_{ij} \sim \text{group} \times \text{time}$
- Unstructured within-subject covariance
- Use `nlme::gls()` with `corSymm` + `varIdent`

---

## Read and Prepare Data

```{r}
#| echo: true
#| output-location: slide
library(dplyr); library(tidyr); library(nlme); library(ggplot2)
if (requireNamespace("emmeans", quietly = TRUE)) library(emmeans)

# Path to data file
path <- "../../data/tlc.csv"

# Read and rename columns
tlc_wide <- read.csv(path, check.names = FALSE, stringsAsFactors = FALSE)
names(tlc_wide) <- trimws(names(tlc_wide))

tlc_wide <- tlc_wide %>%
  rename(
    id  = ID,
    trt = `Treatment Group`,
    w0  = `Lead Level Week 0`,
    w1  = `Lead Level Week 1`,
    w4  = `Lead Level Week 4`,
    w6  = `Lead Level Week 6`
  ) %>%
  mutate(trt = factor(trt, levels = c("P","A"), labels = c("Placebo","Succimer")))

head(tlc_wide)
```

---

## Reshape: Wide to Long

```{r}
#| echo: true
#| output-location: slide
tlc_long <- tlc_wide %>%
  pivot_longer(cols = c(w0, w1, w4, w6),
               names_to = "wk", values_to = "y") %>%
  mutate(
    time = recode(wk, w0 = "0", w1 = "1", w4 = "4", w6 = "6"),
    time = factor(time, levels = c("0","1","4","6")),
    group = trt
  ) %>%
  arrange(id, time) %>%
  select(id, group, time, y)

head(tlc_long)
```

---

## How `gls()` Builds an Unstructured Covariance

::: {.callout-tip}
## UN = `corSymm` + `varIdent`

`nlme` has no single "UN" option. You build it from two pieces:

- `corSymm(form = ~ as.numeric(time) | id)`: **free correlations** among occasions (the off-diagonal of the correlation matrix, all estimated separately).
- `varIdent(form = ~ 1 | time)`: a **separate variance per occasion** (heteroscedastic diagonal).

Together these give a fully unstructured $\Sigma$. The `as.numeric(time) | id` part sets the **within-subject ordering**: it tells `gls()` which occasion each row is and that rows sharing an `id` form one subject's vector. Omit it and the correlations are mis-aligned.
:::

---

## Fit Response Profiles (UN Covariance)

```{r}
#| echo: true
#| output-location: slide
# Unstructured covariance = free correlations (corSymm) + per-occasion variances (varIdent)
fit_un_REML <- gls(
  y ~ group * time,
  data = tlc_long,
  correlation = corSymm(form = ~ as.numeric(time) | id),  # free correlations; sets within-id order
  weights     = varIdent(form = ~ 1 | time),               # one variance per occasion
  method = "REML",
  na.action = na.omit
)

summary(fit_un_REML)
```

::: {.notes}
- This is the central fit of the lecture. Read the model line out loud: group times time gives the saturated cell-means mean model.
- Stress the two-piece UN recipe again: corSymm = free correlations, varIdent = per-occasion variances. nlme has no one-word UN.
- The `as.numeric(time) | id` is the most error-prone piece: it sets the within-subject ordering. If students reuse this code on their own data, they must point it at their own time/id variables.
- Common student error: forgetting `varIdent`, which silently imposes equal variances (compound-symmetry-like), not UN.
- Use REML for the covariance estimate; switch to ML for the fixed-effect LRT on the next slide.
- Transition: now test parallelism by comparing the full and no-interaction models.
:::

---

## Test Group x Time (Parallelism)

```{r}
#| echo: true
#| output-location: slide
# Use ML for fixed-effect LRTs
fit_full_ML   <- update(fit_un_REML, method = "ML")
fit_no_int_ML <- update(fit_full_ML, . ~ group + time)

anova(fit_no_int_ML, fit_full_ML)  # df = (G-1)*(n-1)
```

---

## Interpreting the Parallelism Test

**In plain language:** The likelihood ratio test compares:

- **Full model:** Separate mean for each group-by-time combination (allows non-parallel profiles)
- **Reduced model:** Main effects only (forces parallel profiles)

A significant p-value (e.g., p < 0.05) indicates that profiles are **not parallel** - the treatment effect differs across time points. In the TLC trial, this would mean Succimer's effect is not constant over time (as suggested by the sharp drop then rebound).

---

## Estimated Response Profiles

```{r}
#| echo: true
#| output-location: slide
if (requireNamespace("emmeans", quietly = TRUE)) {
  emm <- emmeans::emmeans(fit_un_REML, ~ group * time)
  emm_df <- as.data.frame(emm)

  print(ggplot(emm_df, aes(x = time, y = emmean, group = group, shape = group)) +
    geom_line() + geom_point(size = 2) +
    labs(y = "Estimated mean blood lead (µg/dL)",
         title = "TLC Trial: Estimated Response Profiles (UN covariance)") +
    theme_minimal())
} else {
  # Alternative: compute fitted values from the model
  newdata <- expand.grid(group = levels(tlc_long$group),
                         time = levels(tlc_long$time))
  newdata$emmean <- predict(fit_un_REML, newdata = newdata)

  print(ggplot(newdata, aes(x = time, y = emmean, group = group, shape = group)) +
    geom_line() + geom_point(size = 2) +
    labs(y = "Estimated mean blood lead (µg/dL)",
         title = "TLC Trial: Estimated Response Profiles (UN covariance)") +
    theme_minimal())
}
```

---

## Reading Off the Covariance: the $\Sigma_i = \mathbf{S}\,\mathbf{C}\,\mathbf{S}$ Identity

`gls()` stores the covariance in two pieces, so we reassemble it from:

$$
\Sigma_i = \mathbf{S}\,\mathbf{C}\,\mathbf{S}
$$

- $\mathbf{S}$ = diagonal of **occasion-specific standard deviations** (from `varIdent`).
- $\mathbf{C}$ = the **correlation matrix** among occasions (from `corSymm`).

(We use $\mathbf{S}$/$\mathbf{C}$ for the SD-correlation split here so as not to collide with the reserved $G$ = random-effects covariance and $R_i$ = residual covariance of Ch. 8.)

So a per-occasion SD on the diagonal times the free correlations rebuilds the full unstructured $\Sigma_i$.

---

## Extract UN Covariance Matrix (advanced / reference)

```{r}
#| echo: true
#| output-location: slide
# --- S: occasion-specific SDs (from varIdent) ---
lev <- levels(tlc_long$time)
# varIdent reports a multiplier for each NON-reference occasion; the
# reference occasion (here "0") is fixed at 1 and is NOT in this vector.
vmult <- coef(fit_un_REML$modelStruct$varStruct, unconstrained = FALSE)
sd_mult <- setNames(rep(1, length(lev)), lev)  # start all at the reference value 1
sd_mult[names(vmult)] <- vmult                 # fill in the estimated multipliers by NAME
SDdiag <- diag(fit_un_REML$sigma * sd_mult)    # S = base sigma * per-occasion multiplier

# --- C: correlation among occasions (from corSymm) ---
Rhat <- corMatrix(fit_un_REML$modelStruct$corStruct)[[1]]

# --- Sigma = S C S ---
Sigma_hat <- SDdiag %*% Rhat %*% SDdiag
dimnames(Sigma_hat) <- list(lev, lev)
round(Sigma_hat, 2)
```

::: {.callout-warning}
## Order-sensitivity (the part that bites)
`varIdent` omits the reference occasion, so you must insert its multiplier of 1 **by name**, not by position; otherwise the reference occasion becomes `NA`. Likewise index everything by `levels(time)` so $\mathbf{S}$ and $\mathbf{C}$ share the same occasion order before multiplying. A positional reindex silently scrambles the rows.
:::

---

## Replicate Figure 5.1: Raw Group Means

```{r}
#| echo: true
#| output-location: slide
means <- tlc_long %>%
  group_by(group, time) %>%
  summarise(mean_y = mean(y, na.rm = TRUE), .groups = "drop") %>%
  mutate(week = as.numeric(as.character(time)))

ggplot(means, aes(x = week, y = mean_y, group = group, shape = group)) +
  geom_line() + geom_point(size = 2) +
  scale_x_continuous(breaks = c(0,1,4,6)) +
  labs(x = "Time (weeks)", y = "Mean blood lead (µg/dL)",
       title = "TLC Trial: Mean Blood Lead by Group and Time") +
  theme_minimal()
```

---

## Baseline-Adjusted ANCOVA (Section 5.6)

Subject-level model:

$$
Y_i^* = \frac{w1_i + w4_i + w6_i}{3}
$$

$$
Y_i^* = \alpha + \beta \cdot \text{group}_i + \gamma \cdot w0_i + \varepsilon_i
$$

---

## ANCOVA Implementation

```{r}
#| echo: true
#| output-location: slide
dat_subj <- tlc_wide %>%
  transmute(
    id,
    group = trt,
    w0, w1, w4, w6,
    Ystar = rowMeans(cbind(w1, w4, w6), na.rm = TRUE)
  )

fit_ancova <- lm(Ystar ~ group + w0, data = dat_subj)
summary(fit_ancova)$coef
```

---

## ANCOVA Output: Interpretation

**In plain language:** The coefficient output shows:

- **Intercept:** Expected average post-baseline lead level when baseline = 0 and group = reference (Placebo)
- **groupSuccimer:** The treatment effect - the difference in average post-baseline lead level for Succimer vs. Placebo, **adjusted for baseline**
- **w0:** The slope relating baseline to follow-up - a positive value means higher baseline is associated with higher follow-up

A negative, significant coefficient for `groupSuccimer` indicates that Succimer reduces blood lead levels compared to Placebo, after accounting for baseline differences.

---

## Same Analysis in SAS: Complete PROC MIXED Program

The R `gls()` fit above corresponds to this standalone SAS program (static, not run here):

```sas
/* TLC response-profiles analysis: unstructured covariance */
proc mixed data=tlc_long method=reml;
    class id group time;
    model y = group time group*time / s chisq;
    repeated time / type=un subject=id r rcorr;
run;
```

- `class` declares the categorical factors (set `id`, `group`, `time` as classification variables).
- `model y = group time group*time` is the saturated cell-means mean model.
- `repeated time / type=un subject=id` is the **unstructured** within-subject covariance (the `corSymm` + `varIdent` pair in R).
- The **`group*time` row of the "Type 3 Tests of Fixed Effects" table is the parallelism test** ($(G-1)(n-1)$ df, here $1\times 3 = 3$ for 2 groups and 4 weeks). The `chisq` option (REML default, no `ddfm=`) reports this as a **Wald chi-square (CHISQ) parallelism test**, matching FLW Table 5.11. It is not a likelihood-ratio test. (No hand-written `contrast` is needed; the Type 3 test already carries the correct df.)
- The R workflow above instead uses an **ML likelihood-ratio test** (`anova()` on two `method="ML"` fits). For an exact SAS analogue of that LRT, refit the full and no-interaction models with `method=ml` and compare $-2\log L$; the Wald chi-square and LR tests target the same hypothesis but are different statistics.

---

## SAS to R Cheat Sheet

| SAS | R |
|-----|---|
| `CLASS id group time;` | Set factor levels (baseline first, Placebo as reference) |
| `MODEL y = group time group*time / S CHISQ;` | `gls(y ~ group * time, ...)` |
| `REPEATED time / TYPE=UN SUBJECT=id;` | `corSymm(~ as.numeric(time) | id) + varIdent(~ 1 | time)` |
| `METHOD=ML` (for LRTs) | `method="ML"` |
| Default REML | `method="REML"` |

---

## Common Mistakes to Avoid

- **Interpreting main effects when interaction is significant**: If parallelism ($H_{01}$) is rejected, main effects for group and time are misleading - the effect depends on the level of the other factor
- **Forgetting to set factor levels correctly**: R's default is to use alphabetical ordering; set the reference level explicitly (e.g., Placebo before Treatment)
- **Using ANCOVA in observational studies without considering Lord's paradox** (defined earlier): with unequal baselines, change-score and ANCOVA conclusions can point in opposite directions
- **Ignoring the relative efficiency of baseline strategies**: Change scores can be substantially less efficient than ANCOVA when baseline-outcome correlation is low
- **Applying response profiles to irregular visit times**: This approach assumes common measurement occasions; use parametric curves (Ch. 6) for irregular timing

---

## Summary

- Response profiles provide a flexible framework for balanced longitudinal data
- Test parallelism ($H_{01}$) before interpreting main effects
- Choose baseline handling strategy based on study design and estimand
- The general linear model (LM) with unstructured covariance maximizes flexibility
- R's `nlme::gls()` replicates SAS PROC MIXED functionality
