---
title: "BIOS 667 - Homework 4"
subtitle: "Chapters 10-13: Residual Diagnostics, Influence Measures, GLMs, Marginal Models & GEE"
author: "Your Name Here"
date: today
format:
  html:
    embed-resources: true
    toc: true
execute:
  warning: false
  message: false
---

## Instructions

**Collaboration Policy:** You may discuss problems with classmates, but all submitted work must be your own.

**AI Use Policy:** AI tools may be used for code assistance. Disclose any AI use in your submission.

**Submission:** Submit a rendered HTML file to Gradescope by the due date.

**Due Date:** See Canvas for due date

## Grading Rubric

| Category | Weight | Description |
|----------|--------|-------------|
| Interpretation & Reasoning | 40% | Conceptual understanding, correct interpretations |
| Analysis & Setup | 40% | Code correctness, model specification |
| Clarity & Presentation | 20% | Writing quality, formatting, organization |

## Setup

```{r}
#| label: setup
#| include: false
library(nlme)
library(lme4)
library(geepack)
library(ggplot2)
library(dplyr)
library(tidyr)
library(MASS)
set.seed(667)
```

---

## Data Description: Epilepsy Clinical Trial

Throughout this homework, we will use data from a placebo-controlled clinical trial of 59 epileptics (Thall and Vail, 1990). Patients with partial seizures were randomized to receive either progabide or placebo as an adjuvant to standard anti-epileptic chemotherapy. Seizure counts were recorded during the 8-week baseline period and during four subsequent 2-week intervals.

```{r}
#| label: load-data
# Load and prepare the epilepsy data with download fallback
data_url <- "https://content.sph.harvard.edu/fitzmaur/ala2e/epilepsy.txt"
data_file <- if (file.exists("../../data/epilepsy.txt")) "../../data/epilepsy.txt" else "data/epilepsy.txt"
if (!file.exists(data_file)) {
  dir.create("data", showWarnings = FALSE)
  download.file(data_url, data_file)
}

epilepsy_raw <- read.table(
  data_file,
  skip = 36,  # Skip the header text
  col.names = c("id", "trt", "age", "baseline", "y1", "y2", "y3", "y4")
)

# Convert to long format for longitudinal analysis
epilepsy <- epilepsy_raw %>%
  pivot_longer(cols = y1:y4, names_to = "visit", values_to = "seizures") %>%
  mutate(
    visit_num = as.numeric(gsub("y", "", visit)),
    trt = factor(trt, labels = c("Placebo", "Progabide")),
    log_baseline = log(baseline / 4 + 0.5),  # log rate per 2 weeks
    log_age = log(age)
  ) %>%
  arrange(id, visit_num)

# Quick look at the data structure
head(epilepsy, 12)
```

---

## Question 1: Residual Diagnostics for Linear Mixed Models (20 points)

For this question, we will first fit a linear mixed-effects model to the log-transformed seizure counts to explore residual diagnostics concepts.

```{r}
#| label: q1-data-prep
# Create log-transformed outcome (add 0.5 to handle zeros)
epilepsy$log_seizures <- log(epilepsy$seizures + 0.5)
```

### Part (a) - **WRITE:** Marginal vs. Conditional Residuals (5 pts)

Explain the difference between **marginal residuals** and **conditional residuals** in the context of linear mixed-effects models. When would you use each type of residual for diagnostic purposes?

*Your answer here*

### Part (b) - **CODE:** Fit LMM and Extract Residuals (5 pts)

Fit a random intercept model predicting `log_seizures` from `visit_num`, `trt`, their interaction, and `log_baseline`. Extract both marginal and conditional residuals.

```{r}
#| label: q1b-solution
# Your code here
```

### Part (c) - **WRITE:** Interpret Residual Plots (5 pts)

Examine the residual plots below and describe what each plot reveals about model adequacy. Are there any concerns?

```{r}
#| label: q1c-plots
#| fig-height: 8
#| fig-width: 10
# Your code here to create residual diagnostic plots
# Include: residuals vs fitted, residuals vs time, QQ plot, histogram
```

*Your answer here*

### Part (d) - **CODE:** Check Random Effects Normality (5 pts)

Extract the random effects (BLUPs) and create a QQ plot to assess normality. Interpret the results.

```{r}
#| label: q1d-solution
#| fig-height: 5
#| fig-width: 6
# Your code here
```

*Your answer here*

---

## Question 2: Influence Diagnostics for Mixed Models (20 points)

### Part (a) - **WRITE:** Level-1 vs. Level-2 Outliers (5 pts)

In mixed models, we distinguish between **Level-1 outliers** (unusual observations) and **Level-2 outliers** (unusual clusters/subjects). Explain this distinction and why cluster-level influence diagnostics are particularly important in longitudinal data analysis.

*Your answer here*

### Part (b) - **CODE:** Cook's Distance-Like Measure (5 pts)

Implement a cluster deletion diagnostic by refitting the model without each subject and computing the change in fixed effect estimates. Identify the most influential subjects.

::: {.callout-tip collapse="true"}
## Hint: Cluster Deletion Diagnostic Algorithm

This involves iteratively removing each subject and measuring the impact on parameter estimates. Here is the general approach:

**Pseudocode:**
```
1. Fit the full model to all data and extract fixed effects (beta_full)
2. Get list of unique subject IDs
3. Create empty storage for results (e.g., a matrix)
4. For each subject i in the data:
   a. Create dataset excluding subject i
   b. Refit the model on this reduced dataset
   c. Extract fixed effects (beta_minus_i)
   d. Calculate change: delta_i = beta_full - beta_minus_i
   e. Store the change (or a summary like Cook's D-like statistic)
5. Identify subjects with largest influence
```

**R code structure:**
```r
# Get unique subject IDs
subjects <- unique(epilepsy$id)

# Storage for influence measures
influence_results <- data.frame(
  id = subjects,
  delta_intercept = NA,
  delta_trt = NA,
  cooks_d = NA
)

# Loop over subjects
for (i in seq_along(subjects)) {
  # Exclude subject i
  data_minus_i <- epilepsy[epilepsy$id != subjects[i], ]

  # Refit model (use try() to handle any fitting issues)
  fit_minus_i <- try(lme(...), silent = TRUE)

  if (!inherits(fit_minus_i, "try-error")) {
    # Calculate influence measures
    # Store results...
  }
}
```

**Note**: This can take some time to run since you are refitting the model ~59 times.
:::

```{r}
#| label: q2b-solution
# Your code here
```

### Part (c) - **WRITE:** Interpret Influential Subjects (5 pts)

Examine the data for the 2-3 most influential subjects identified above. What characteristics make them influential?

```{r}
#| label: q2c-examination
# Your code here to examine the most influential subjects
```

*Your answer here*

### Part (d) - **CODE:** Sensitivity Analysis (5 pts)

Refit the model excluding the most influential subject and compare the fixed effect estimates to the full model. Discuss whether conclusions change.

```{r}
#| label: q2d-solution
# Your code here
```

*Your answer here*

---

## Question 3: Generalized Linear Models Review and Marginal Models (20 points)

### Part (a) - **WRITE:** Link Functions and Variance Functions (5 pts)

For count data like seizure counts, explain why a **Poisson regression** model might be appropriate. Specify the:

1. Canonical link function
2. Variance function
3. Mean-variance relationship

Also discuss what **overdispersion** means and why it is a concern for the epilepsy data.

*Your answer here*

### Part (b) - **CODE:** Fit Marginal Model with GEE (5 pts)

Fit a marginal model using GEE for the seizure count data with a Poisson family, log link, and **exchangeable** working correlation structure. Include `visit_num`, `trt`, their interaction, and `log_baseline` as predictors.

```{r}
#| label: q3b-solution
# Your code here
```

### Part (c) - **WRITE:** Interpret GEE Coefficients (5 pts)

Interpret the following coefficients from your GEE model in terms of seizure rates:

1. The treatment main effect (`trtProgabide`)
2. The treatment-by-time interaction (`visit_num:trtProgabide`)
3. The baseline effect (`log_baseline`)

```{r}
#| label: q3c-coefficients
# Your code here to extract and display coefficients
```

*Your answer here*

### Part (d) - **CODE:** Compare Working Correlation Structures (5 pts)

Fit GEE models with **independence**, **AR(1)**, and **unstructured** working correlation structures. Compare the estimated coefficients and robust standard errors across models.

```{r}
#| label: q3d-solution
# Your code here
```

*Your answer here*

---

## Question 4: Robust Standard Errors and GEE Inference (20 points)

### Part (a) - **WRITE:** Sandwich Estimator Derivation (5 pts)

The robust (sandwich) variance estimator for GEE is:
$$\widehat{\text{Var}}(\hat{\beta}) = B_0^{-1} B_1 B_0^{-1}$$

Explain what each component represents:

1. $B_0$ (the "bread")
2. $B_1$ (the "meat")

Why is this estimator called "robust" and what is it robust to?

*Your answer here*

### Part (b) - **CODE:** Model-Based vs. Robust Standard Errors (5 pts)

For the exchangeable GEE model from Question 3, extract and compare the model-based (naive) and robust standard errors. Discuss the implications of any differences.

::: {.callout-tip collapse="true"}
## Hint: Extracting Standard Errors from geepack

The `geepack::geeglm()` function stores both naive and robust variance estimates. After fitting a model called `gee_fit`, you can access them via:

```r
# Summary gives robust SEs by default
summary(gee_fit)

# For naive (model-based) SEs, access the geese object:
# Naive variance-covariance matrix
naive_vcov <- gee_fit$geese$vbeta.naiv

# Robust variance-covariance matrix
robust_vcov <- gee_fit$geese$vbeta

# Convert to standard errors
naive_se <- sqrt(diag(naive_vcov))
robust_se <- sqrt(diag(robust_vcov))
```
:::

```{r}
#| label: q4b-solution
# Your code here
```

*Your answer here*

### Part (c) - **WRITE:** Small Sample Corrections (5 pts)

With only 59 subjects in the epilepsy study, small sample bias in the sandwich estimator may be a concern. Describe two approaches to address this issue and explain when they might be preferred.

*Your answer here*

### Part (d) - **CODE:** Apply Small Sample Correction (5 pts)

Apply a small sample correction to the GEE analysis using an appropriate R package or manual implementation. Compare the corrected and uncorrected results.

```{r}
#| label: q4d-solution
# Your code here
```

*Your answer here*

---

## Question 5: Model Selection and Comprehensive Analysis (20 points)

### Part (a) - **WRITE:** GEE vs. GLMM (5 pts)

Compare and contrast **GEE (marginal models)** and **GLMMs (conditional models)** for analyzing the epilepsy data. Address:

1. Interpretation of coefficients
2. Assumptions required
3. When each approach is preferred

*Your answer here*

### Part (b) - **CODE:** Fit GLMM for Comparison (5 pts)

Fit a Poisson GLMM with random intercepts using `lme4::glmer()`. Compare the fixed effect estimates to the GEE results and explain any differences.

```{r}
#| label: q5b-solution
# Your code here
```

*Your answer here*

### Part (c) - **CODE:** Model Diagnostics for GLMM (5 pts)

Assess the GLMM by examining:

1. Deviance residuals
2. Overdispersion
3. Random effects distribution

```{r}
#| label: q5c-solution
#| fig-height: 10
#| fig-width: 10
# Your code here
```

*Your answer here*

### Part (d) - **CODE & WRITE:** Comprehensive Analysis Summary (5 pts)

Write a brief results paragraph (as you would for a journal article) summarizing the treatment effect from either the GEE or GLMM analysis. Include:

1. Effect estimate with confidence interval
2. Statistical significance
3. Clinical interpretation
4. Any caveats or limitations

```{r}
#| label: q5d-summary
# Your code here to generate summary statistics
```

*Your answer here*

---

## Peer Review Section

After submission, you will be assigned peer reviews. Use this rubric:

| Criterion | Excellent (5) | Good (3-4) | Needs Work (1-2) |
|-----------|---------------|------------|------------------|
| Code correctness | All code runs, correct output | Minor errors | Major errors |
| Interpretations | Clear, accurate, in context | Mostly correct | Missing key points |
| Presentation | Well-organized, readable | Adequate | Difficult to follow |

---

## Reminders

- [ ] Knit document before submission
- [ ] Check that all code executes without errors
- [ ] Ensure figures are visible and properly labeled
- [ ] Format tables appropriately
- [ ] Include AI disclosure if applicable
